---
title: "small_species_ext_mort"
author: "Elyza Pilipaityte"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

clear memory

```{r}
rm(list=ls())

```

Install mizer, load libraries

```{r}
# remotes::install_github("sizespectrum/mizerExperimental", ref = "tuneMR_w_max", force=T)
remotes::install_github("sizespectrum/mizerExperimental")
# remotes::install_github("sizespectrum/mizerMR")

library(mizerMR)
library(tidyverse)
library(assertthat)
library(mizerHowTo)
library(patchwork)
```

```{r}
# Define my own plotting 'theme' (called 'My_theme')
My_theme <- theme(
  panel.background = element_blank(),
  axis.line = element_line(linewidth = 1),  # draws x and y axis lines
  text = element_text(size = 17, family = "gillsans")
)
```

For this model, we had the following main criteria:

1)  Modelled biomasses were within 20% of the average observed biomasses from 1992-2001.
2)  The biomass size spectrum slope was negative and close to -1.
3)  Growth curves match observed size-at-age data for each species.
4)  Perch and pikeperch diets captured dietary changes with body size and became more piscivorous at larger sizes.
5)  Recruitment parameters ensured single-species yield curves were dome-shaped, as expected by theory.
6)  The modeled catches through time captured the trends in the reported catches.

### Setting mizer model

Load parameters

```{r warning=F, message=FALSE, echo=F, eval = T}
# Read in species parameters, gear parameters and the interaction matrix files
curonian_params <- read.csv("curonian_params3.csv")
curonian_interaction <- read.csv("curonian_interaction2.csv") 
curonian_gear_params <- read.csv("curonian_gear_params.csv")

```

```{r warning=F, message=FALSE, echo=F, eval = T}
# Modify gear names to match species names
curonian_gear_params$gear <- curonian_gear_params$species

# effort vector
initial_effort <- c(0, 0, 0.111, 0.006, 0.131, 0.077, 0.083)

cur_model <- newMultispeciesParams(species_params = curonian_params, 
                                   interaction = curonian_interaction,
                                   gear_params = curonian_gear_params,
                                   initial_effort = initial_effort, no_w = 200,
                                   lambda = 2.05, n = 3/4, p = 0.7,
                                   resource_rate = 10)

cur_model <- steady(cur_model, t_max = 200)
plotSpectra(cur_model, power = 2, total = TRUE)

```

# Setting allometric external death rate

```{r}
# Setting coefficient for each species
z0pre <- c("small_fish" = 0.1, "ruffe" = 0.1, "bream" = 0.1, "white_bream" = 0.1, "roach" = 0.1, "perch" = 0.1, "pikeperch" = 0.1)
# Multiply by power of size with exponent, here chosen to be -1/4
# The outer() function makes it an array species x size
allo_mort <- outer(z0pre, w(cur_model)^(-1/4))
# Change the external mortality rate in the params object
ext_mort(cur_model) <- allo_mort
```

### Biomass, growth and yields calibration

Defining function

```{r}
check_convergence <- function(model, params, tol_biomass = 0.2, tol_age = 0.5, tol_yield = 0.2) {
  model_biomass <- getBiomass(model)
  obs_biomass <- params$biomass_observed
  
  # Biomass convergence
  biomass_diff <- abs(model_biomass - obs_biomass)
  biomass_converged <- all(biomass_diff < tol_biomass)
  
  # Age convergence
  age_diff <- abs(age_mat(model) - params$age_mat)
  age_converged <- all(age_diff < tol_age)
  
  # Yield convergence (only for species with observed data)
  yield_converged <- TRUE
  max_yield_diff <- 0
  
  if(!is.null(params$yield_observed)) {
    model_yield <- getYield(model)
    
    # Only checking species that have observed yield data (not NA)
    valid_species <- !is.na(params$yield_observed)
    
    if(any(valid_species)) {
      yield_diff <- abs(model_yield[valid_species] - params$yield_observed[valid_species])
      yield_converged <- all(yield_diff < tol_yield)
      max_yield_diff <- max(yield_diff)
    }
  }
  
  list(converged = biomass_converged & age_converged & yield_converged,
       biomass_converged = biomass_converged,
       age_converged = age_converged,
       yield_converged = yield_converged,
       max_biomass_diff = max(biomass_diff),
       max_age_diff = max(age_diff),
       max_yield_diff = max_yield_diff)
}
```

Calibration

```{r}
# Iterative calibration
cur_model <- steady(cur_model, t_max = 100)

max_iter <- 50

for(i in 1:max_iter) {
  cur_model <- cur_model |>
    calibrateBiomass() |>
    matchBiomasses() |>
    steady(t_max = 100) |>
    matchGrowth() |>
    steady(t_max = 100)
  
  # yield matching only for species with observed yield data
  if(!is.null(curonian_params$yield_observed) && any(!is.na(curonian_params$yield_observed))) {
    
    cur_model <- cur_model |>
      matchYield() |>  
      steady(t_max = 100)
  }
  
  conv <- check_convergence(cur_model, curonian_params, 
                           tol_biomass = 0.2, tol_age = 0.5, tol_yield = 0.2)
  
  if(i %% 5 == 0) {
    yield_status <- if(is.na(conv$max_yield_diff)) "NA" else sprintf("%.2f", conv$max_yield_diff)
    cat(sprintf("Iter %d - Biomass: %s (%.2f), Age: %s (%.2f yr), Yield: %s (%s)\n",
                i, conv$biomass_converged, conv$max_biomass_diff,
                conv$age_converged, conv$max_age_diff,
                conv$yield_converged, yield_status))
  }
  
  if(conv$converged) {
    cat("Converged after", i, "iterations\n")
    break
  }
}

cur_model@catchability
```

Resource abundance (kappa) calibration

```{r}
cur_model@resource_params$kappa # very low now

cur_model <- scaleDownBackground(cur_model, factor = 1/2)
```

Running calibration again after rescaling resource abundance (repeated several times to get desirable kappa value 6 or 7)

### Tune resilience

If I set erepro to 0.001 it converges faster

```{r}
# I change erepro for all species to 0.001, R_max is also changed when using setBevertonHolt()
cur_model <- setBevertonHolt(cur_model, erepro = 0.001)
species_params(cur_model) |> select(erepro, R_max)
# Let’s see what reproduction levels this gives:
getReproductionLevel(cur_model)
# the reproduction level is the ratio between RDD and R_max and can vary between 0 and 1. It tells us how close the actual reproduction (after applying density dependence) is to the theoretical maximum, set by R_max
getRDD(cur_model) / species_params(cur_model)$R_max
# how close the density dependent reproduction rate RDD is to the density independent reproduction rate RDI:
getRDI(cur_model) / getRDD(cur_model)
```

Tuning reproduction level (this code chunk will take around 30 minutes to run if I start from erepro 0.001 if from initial calibrated erepro, then it can take several hours)

```{r}
rep_level <- getReproductionLevel(cur_model)

# Safety bounds for reproduction levels
RL_MIN <- 0.1
RL_MAX <- 0.999

cat("Checking initial reproduction levels...\n")
print(rep_level)

if(any(rep_level < RL_MIN | rep_level > RL_MAX)) {
  cat("\nCorrecting out-of-bounds reproduction levels:\n")
  problematic <- rep_level < RL_MIN | rep_level > RL_MAX
  print(rep_level[problematic])
  
  rep_level[rep_level < RL_MIN] <- 0.1
  rep_level[rep_level > RL_MAX] <- RL_MAX
  cur_model <- setBevertonHolt(cur_model, reproduction_level = rep_level)
  
  cat("Corrected values:\n")
  print(rep_level[problematic])
}

sppList <- cur_model@species_params$species

# Fmsy target ranges
FlevelMax <- c(0.5, 0.5, 0.7, 0.7, 0.7, 0.5, 0.3)  # Bream, Perch, Pikeperch, Roach, Ruffe, Small Fish, White Bream
FlevelMin <- c(0.1, 0.1, 0.05, 0.1, 0.1, 0.1, 0.1)

# F crash targets - where yield should be close to zero
# Species order: Bream, Perch, Pikeperch, Roach, Ruffe, Small Fish, White Bream
FcrashTarget <- c(1.0, 1.1, 1.5, 1.5, 1.5, 1.1, 1.0)
FcrashTolerance <- 0.15  # Allow ±0.15 around target

test_value <- rep(NA, 7)
crash_F <- rep(NA, 7)
tracking_df <- data.frame()

for (i in 1:100) {
  cat(sprintf("\n========== Iteration %d ==========\n", i))
  
  # Batch update approach
  rep_level_new <- getReproductionLevel(cur_model)
  
  for (n in 1:7) {
    current_species <- sppList[n]
    currentRL <- rep_level_new[current_species]
    
    # Extended F range to capture crash point
    if(i > 1 && !is.na(test_value[n])) {
      F_center <- test_value[n]
      F_range <- seq(max(0.05, F_center - 0.4), min(3.0, F_center + 1.5), 0.05)
    } else {
      F_range <- seq(0.05, 3.0, 0.05)  # Extended to F=3.0
    }
    
    temp <- getYieldVsF(cur_model, species = current_species, F_range = F_range)
    
    # CRASH DETECTION: Yield close to zero (absolute threshold)
    # "Close to zero" means yield < 0.01 (1% of a typical yield of 1.0)
    yield_threshold <- 0.01  # Absolute threshold - yield must be nearly zero
    
    # Finding where population crashes (yield becomes nearly zero)
    # Looking for first point after Fmsy where yield drops below threshold
    fmsy_idx <- which.max(temp$yield)
    post_fmsy <- temp$F > temp$F[fmsy_idx]
    crash_candidates <- which(temp$yield < yield_threshold & post_fmsy)
    
    if(length(crash_candidates) > 0) {
      crash_F[n] <- temp$F[min(crash_candidates)]
    } else {
      crash_F[n] <- max(F_range)  # No crash detected in range
    }
    
    # Get Fmsy
    test_value[n] <- temp$F[which.max(temp$yield)]
    
    # Calculate distance from crash target
    distance_crash <- abs(crash_F[n] - FcrashTarget[n])
    
    # Adaptive step size based on distance to crash target
    step_size_base <- 0.02
    if(distance_crash > 0.5) {
      step_size <- step_size_base  # Full step when far
    } else if(distance_crash > 0.3) {
      step_size <- step_size_base * 0.75  # 75% step
    } else if(distance_crash > 0.15) {
      step_size <- step_size_base * 0.5  # 50% step - getting close
    } else {
      step_size <- step_size_base * 0.25  # 25% step - very close, fine tune
    }
    
    action <- "no change"
    reason <- ""
    
    # Check both criteria
    fmsy_ok <- test_value[n] >= FlevelMin[n] & test_value[n] <= FlevelMax[n]
    fcrash_ok <- abs(crash_F[n] - FcrashTarget[n]) <= FcrashTolerance
    
    # CORRECTED LOGIC:
    # Higher RL = More resilient = Crashes at HIGHER F
    # Lower RL = Less resilient = Crashes at LOWER F
    
    if(!fcrash_ok) {
      # Crash criterion takes priority
      if(crash_F[n] < FcrashTarget[n]) {
        # Crashes too early -> need MORE resilience -> INCREASE RL
        new_rl <- min(RL_MAX, currentRL + step_size)
        rep_level_new[current_species] <- new_rl
        action <- sprintf("↑ %.3f (%.3f→%.3f)", step_size, currentRL, new_rl)
        reason <- "[CRASH EARLY - increasing resilience]"
        
        # Warning if hitting maximum
        if(new_rl == RL_MAX && currentRL == RL_MAX) {
          reason <- paste0(reason, " ⚠️ AT MAXIMUM")
        }
      } else if(crash_F[n] > FcrashTarget[n] + FcrashTolerance) {
        # Crashes too late -> need LESS resilience -> DECREASE RL
        new_rl <- max(RL_MIN, currentRL - step_size)
        rep_level_new[current_species] <- new_rl
        action <- sprintf("↓ %.3f (%.3f→%.3f)", step_size, currentRL, new_rl)
        reason <- "[CRASH LATE - decreasing resilience]"
      }
    } else if(!fmsy_ok) {
      # Crash OK, now tunning Fmsy with smaller steps
      step_size_fmsy <- step_size_base * 0.3
      
      if(test_value[n] > FlevelMax[n]) {
        # Fmsy too high -> need less productivity -> DECREASE RL
        new_rl <- max(RL_MIN, currentRL - step_size_fmsy)
        rep_level_new[current_species] <- new_rl
        action <- sprintf("↓ %.3f (%.3f→%.3f)", step_size_fmsy, currentRL, new_rl)
        reason <- "[Fmsy HIGH]"
      } else if(test_value[n] < FlevelMin[n]) {
        # Fmsy too low -> need more productivity -> INCREASE RL
        new_rl <- min(RL_MAX, currentRL + step_size_fmsy)
        rep_level_new[current_species] <- new_rl
        action <- sprintf("↑ %.3f (%.3f→%.3f)", step_size_fmsy, currentRL, new_rl)
        reason <- "[Fmsy LOW]"
      }
    }
    
    cat(sprintf("  %s: Fmsy=%.3f [%.2f-%.2f]%s | Fcrash=%.2f [~%.2f]%s | %s %s\n",
                current_species, 
                test_value[n], FlevelMin[n], FlevelMax[n], ifelse(fmsy_ok, "✓", "✗"),
                crash_F[n], FcrashTarget[n], ifelse(fcrash_ok, "✓", "✗"),
                action, reason))
    
    # Track
    tracking_df <- rbind(tracking_df, data.frame(
      iteration = i,
      species = current_species,
      Fmsy = test_value[n],
      Fcrash = crash_F[n],
      Fcrash_target = FcrashTarget[n],
      rep_level = rep_level_new[current_species],  # Track NEW value
      fmsy_ok = fmsy_ok,
      fcrash_ok = fcrash_ok,
      both_ok = fmsy_ok & fcrash_ok
    ))
  }
  
  # Apply all changes
  cur_model <- setBevertonHolt(cur_model, reproduction_level = rep_level_new)
  
  # Check convergence (both criteria)
  fmsy_bounds <- test_value >= FlevelMin & test_value <= FlevelMax
  fcrash_bounds <- abs(crash_F - FcrashTarget) <= FcrashTolerance
  both_ok <- fmsy_bounds & fcrash_bounds
  n_converged <- sum(both_ok)
  
  cat(sprintf("\n  → %d/7 species converged (Fmsy: %d/7, Fcrash: %d/7)\n",
              n_converged, sum(fmsy_bounds), sum(fcrash_bounds)))
  
  if (all(both_ok)) {
    cat(sprintf("\n✓ All species converged after %d iterations!\n", i))
    break
  }
  
  # Early exit if stuck
  if(i > 30) {
    prev_converged <- sum(tracking_df$both_ok[tracking_df$iteration == max(1, i-15)])
    if(n_converged == prev_converged) {
      cat("\nWarning: No progress in last 15 iterations. May need manual adjustment.\n")
      cat("Consider adjusting FcrashTolerance or step sizes.\n")
      break
    }
  }
}

# Summary
final <- data.frame(
  species = sppList,
  Fmsy = test_value,
  Fmsy_min = FlevelMin,
  Fmsy_max = FlevelMax,
  Fcrash = crash_F,
  Fcrash_target = FcrashTarget,
  Fcrash_diff = crash_F - FcrashTarget,
  rep_level = getReproductionLevel(cur_model),
  Fmsy_ok = test_value >= FlevelMin & test_value <= FlevelMax,
  Fcrash_ok = abs(crash_F - FcrashTarget) <= FcrashTolerance,
  converged = (test_value >= FlevelMin & test_value <= FlevelMax) & 
              (abs(crash_F - FcrashTarget) <= FcrashTolerance)
)

cat("\n========== FINAL RESULTS ==========\n")
print(final)
```

### time series yields

I will check how well our modeled yields match with observed time series yields

```{r}
# Re-read with Year column
obsy <- read.csv("obslandings.csv")

# If Year column exists, just rename
colnames(obsy) <- c("time", "sp", "value")

# Fix species names
obsy$sp <- tolower(obsy$sp)
obsy$sp[obsy$sp == "silverbream"] <- "white_bream"
```

```{r}
# Set up gears (one per species)
sim1 <- project(cur_model, t_max = 300, dt = 0.2, effort = initial_effort)

current_gear <- gear_params(sim1@params)
current_gear$gear <- current_gear$species
gear_params(sim1@params) <- current_gear
sim1@params <- setParams(sim1@params)


# Read F data with confidence intervals
f_data <- read.csv("CLF_1990_2023matrix_ci.csv")  

years_data <- f_data$Year  

# Create matrices
main_species <- c("Bream", "Perch", "Pikeperch", "Roach", "White_bream")
f_main <- as.matrix(f_data[, main_species])
f_low <- as.matrix(f_data[, paste0(main_species, "_low")])  
f_up <- as.matrix(f_data[, paste0(main_species, "_up")])

# Set dimnames properly
rownames(f_main) <- as.character(years_data)
rownames(f_low) <- as.character(years_data)
rownames(f_up) <- as.character(years_data)
colnames(f_main) <- tolower(main_species)
colnames(f_low) <- tolower(main_species)
colnames(f_up) <- tolower(main_species)

# Adding missing species
f_main_complete <- cbind(small_fish = rep(0, nrow(f_main)), 
                         ruffe = rep(0, nrow(f_main)), 
                         f_main)
f_low_complete <- cbind(small_fish = rep(0, nrow(f_low)), 
                        ruffe = rep(0, nrow(f_low)), 
                        f_low)
f_up_complete <- cbind(small_fish = rep(0, nrow(f_up)), 
                       ruffe = rep(0, nrow(f_up)), 
                       f_up)

# Setting rownames for complete matrices
rownames(f_main_complete) <- as.character(years_data)
rownames(f_low_complete) <- as.character(years_data)
rownames(f_up_complete) <- as.character(years_data)

# 2. Run to steady state
sim2 <- project(sim1@params, effort = initial_effort, t_max = 300, dt = 0.1,
                initial_n = sim1@n[300,,], initial_npp = sim1@n_pp[300,])

# Reordering and continuing as before
species_order <- species_params(sim2@params)$species
f_main_ordered <- f_main_complete[, species_order]
f_low_ordered <- f_low_complete[, species_order]
f_up_ordered <- f_up_complete[, species_order]

# Converting to effort
relative_effort <- sweep(f_main_ordered, 2, gear_params(sim2@params)$catchability, "/")
relative_effort_low <- sweep(f_low_ordered, 2, gear_params(sim2@params)$catchability, "/")
relative_effort_up <- sweep(f_up_ordered, 2, gear_params(sim2@params)$catchability, "/")
# exclude small fish and ruffe
# Set NaN values to 0 but keep all species
relative_effort[is.nan(relative_effort)] <- 0
relative_effort_low[is.nan(relative_effort_low)] <- 0
relative_effort_up[is.nan(relative_effort_up)] <- 0


# Add ramp-up with fixed function
# Define the add_ramp_up function
add_ramp_up <- function(effort_matrix) {
  firstRecordedEffort <- effort_matrix[1,]
  
  # Create numeric time sequence
  ramp_years <- 1893:1991
  data_years <- as.numeric(rownames(effort_matrix))
  
  initial_effort1 <- matrix(0, nrow = length(ramp_years), ncol = ncol(effort_matrix),
                           dimnames = list(as.character(ramp_years), colnames(effort_matrix)))
  
  for (i in 1:ncol(effort_matrix)) {
    initial_effort1[, i] <- seq(from = 0, to = firstRecordedEffort[i], 
                               length.out = nrow(initial_effort1))
  }
  
  # Ensure rownames are character (will be converted to numeric by mizer)
  combined_effort <- rbind(initial_effort1, effort_matrix)
  rownames(combined_effort) <- as.character(c(ramp_years, data_years))
  
  return(combined_effort)
}

# Now I can use it
relative_effort_full <- add_ramp_up(relative_effort)
relative_effort_low_full <- add_ramp_up(relative_effort_low)
relative_effort_up_full <- add_ramp_up(relative_effort_up)

# Running simulations
sim3_main <- project(sim2@params, effort = relative_effort_full, 
                     dt = 0.1, t_save = 1,
                     initial_n = sim2@n[300,,], 
                     initial_npp = sim2@n_pp[300,])

sim3_low <- project(sim2@params, effort = relative_effort_low_full, 
                    dt = 0.1, t_save = 1,
                    initial_n = sim2@n[300,,], 
                    initial_npp = sim2@n_pp[300,])

sim3_up <- project(sim2@params, effort = relative_effort_up_full, 
                   dt = 0.1, t_save = 1,
                   initial_n = sim2@n[300,,], 
                   initial_npp = sim2@n_pp[300,])

# Preparing yield data
y_main <- getYield(sim3_main) %>% melt() %>% mutate(scenario = "main")
y_low <- getYield(sim3_low) %>% melt() %>% mutate(scenario = "low")
y_up <- getYield(sim3_up) %>% melt() %>% mutate(scenario = "up")

# Combine all scenarios
y_all <- rbind(y_main, y_low, y_up)
y_wide <- reshape(y_all, idvar = c("time", "sp"), timevar = "scenario", direction = "wide")
colnames(y_wide) <- c("time", "sp", "main", "low", "up")

# Create plot with ribbon
ggplot(y_wide) + 
  geom_ribbon(aes(x = time, ymin = low, ymax = up, fill = sp), alpha = 0.3) +
  geom_line(aes(x = time, y = main, colour = sp), size = 1) +
  geom_point(data = obsy, aes(x = time, y = value, colour = sp), size = 2) +
  facet_wrap(~sp, scales = "free_y") +
  scale_y_continuous(name = "Yield [g/m-2/year]") +
  xlim(1992, 2021) +
  labs(title = "Model yield with F confidence intervals vs observations") +
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend since facets show species
```

plot used in supplements

```{r}
# Filter out small fish and ruffe
y_wide_filtered <- y_wide %>%
  filter(!sp %in% c("small_fish", "ruffe"))
obsy_filtered <- obsy %>%
  filter(!sp %in% c("small_fish", "ruffe"))

# Set panel order
sp_order <- c("bream", "white_bream", "roach", "perch", "pikeperch")  
y_wide_filtered$sp <- factor(y_wide_filtered$sp, levels = sp_order)
obsy_filtered$sp   <- factor(obsy_filtered$sp,   levels = sp_order)

# Calculate Spearman correlation (filtered data)
r_square <- y_wide_filtered %>%
  inner_join(obsy_filtered, by = c("time", "sp")) %>%
  filter(!is.na(main), !is.na(value)) %>%
  group_by(sp) %>%
  summarise(
    spearman = cor(main, value, method = "spearman"),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("ρ = ", round(spearman, 3)),
    time = 2018,
    y = Inf
  )

# Model yield with F confidence intervals vs observations
timeyield <- ggplot(y_wide_filtered) + 
  geom_ribbon(aes(x = time, ymin = low, ymax = up), alpha = 0.3) +
  geom_line(aes(x = time, y = main), linewidth = 1) +
  geom_hline(yintercept = -Inf, linewidth = 0.5) +  
  geom_vline(xintercept = -Inf, linewidth = 0.5) +
  geom_point(data = obsy_filtered, aes(x = time, y = value), size = 2) +
  geom_text(data = r_square, 
            aes(x = time, y = y, label = label), 
            vjust = 1, hjust = 0.9, size = 4, fontface = "bold") +
  facet_wrap(~sp, scales = "free_y", ncol = 2, labeller = as_labeller(function(x) gsub("_", " ", x))) +
  scale_y_continuous(
    name = "Yield (g/m²/year)",
    expand = expansion(mult = c(0.05, 0.15))
  ) +
  xlim(1992, 2021) +
 labs(x = "Time (year)") + 
 theme_classic() +
  theme(
    text = element_text(size = 17, family = "gillsans"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 12, hjust = 0),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.line = element_blank(),
    legend.position = "none"
  )

# ggsave("timeyield.tiff", timeyield, width = 8, height = 7, dpi = 600, bg = "white")
```

### Best models

I ran all calibration steps above and saved 40 different calibrated models with slightly different kappa and external mortality values. I then saved their data — fish percent in predator diets, feeding level of the main predator, and the strength of the correlation between modeled and observed time series — for comparison, in order to evaluate which five models are the best for the analysis.

```{r}
models <- read.csv("models.csv")

# Method 1: Simple Ranking (lower is better)
models$rank_fish <- rank(-models$fish_in_pikeperch_diet)
models$rank_feeding <- rank(-models$Feeding_level_of_pikeperch)
models$rank_corr <- rank(-models$Sum_of_Spearman_corr)
models$rank_mean_corr <- rank(-models$Mean_corr_Spearman)

models$total_rank <- models$rank_fish + models$rank_feeding + models$rank_corr + models$rank_mean_corr

# Method 2: Normalized Scoring (0-1 scale)
normalize <- function(x) {
  (x - min(x)) / (max(x) - min(x))
}

models$norm_fish <- normalize(models$fish_in_pikeperch_diet)
models$norm_feeding <- normalize(models$Feeding_level_of_pikeperch)
models$norm_corr_sum <- normalize(models$Sum_of_Spearman_corr)
models$norm_corr_mean <- normalize(models$Mean_corr_Spearman)

models$avg_normalized <- (models$norm_fish + models$norm_corr_sum) / 2

# Method 3: Weighted Score (equal weights)
models$weighted_equal <- (models$norm_fish + models$norm_feeding + models$norm_corr_sum + models$norm_corr_mean) / 4

# Method 4: Weighted Score (different weights)
# Example: fish_diet=0.4, feeding_level=0.3, correlation=0.3
models$weighted_custom <- models$norm_fish * 0.5 +
                        models$norm_corr_sum * 0.5

# Display results sorted by different methods
cat("\n=== TOP 5 MODELS BY RANKING METHOD ===\n")
print(models[order(models$total_rank), c("Nr", "Kappa", "Ext_mort", "Ext_Mort_slope", "fish_in_pikeperch_diet", "Feeding_level_of_pikeperch", "Sum_of_Spearman_corr", "Mean_corr_Spearman", "total_rank")][1:8,])

cat("\n=== TOP 5 MODELS BY NORMALIZED AVERAGE ===\n")
print(models[order(-models$avg_normalized), c("Nr", "Kappa", "Ext_mort", "Ext_Mort_slope", "fish_in_pikeperch_diet", "Feeding_level_of_pikeperch", "Sum_of_Spearman_corr", "Mean_corr_Spearman", "total_rank")][1:8,])

cat("\n=== TOP 5 MODELS BY WEIGHTED CUSTOM ===\n")
print(models[order(-models$weighted_custom), c("Nr", "Kappa", "Ext_mort", "Ext_Mort_slope", "fish_in_pikeperch_diet", "Feeding_level_of_pikeperch", "Sum_of_Spearman_corr", "Mean_corr_Spearman", "total_rank")][1:8,])

```

Now I will use these five models to show a range in the calibration evaluation plots.

I load all five already-calibrated models to make plots with ranges

```{r}
models <- list(
  M3  = read_rds("M6_kappa8_lambda2.05extMort01_0.333.rds"),
  M4  = read_rds("M4_kappa7_lambda2.05extMort01_0.25.rds"),
  M5  = read_rds("M5_kappa7_lambda2.05extMort01_0.333.rds"),
  M12 = read_rds("M12_kappa7_lambda2.05extMort02_0.333.rds"),
  M13 = read_rds("M7_kappa9_lambda2.05extMort01_0.333.rds")
)
```

### Calibration Results

Model properties exploration

# Yields

```{r}
library(purrr)

species_list <- c(
  "small_fish"  = "small_fish",
  "ruffe"       = "ruffe",
  "bream"       = "bream",
  "white_bream" = "white_bream",
  "roach"       = "roach",
  "perch"       = "perch",
  "pikeperch"   = "pikeperch"
)

# F_range per species 
f_ranges <- list(
  small_fish  = seq(0,    1.5, 0.01),
  ruffe       = seq(0.1,  1.5, 0.01),
  white_bream = seq(0.01, 1.5, 0.01),
  roach       = seq(0.01, 1.5, 0.01),
  perch       = seq(0,    1.5, 0.01),
  bream       = seq(0.01, 1.5, 0.01),
  pikeperch   = seq(0,    1.5, 0.01)
)

plot_data <- imap_dfr(models, function(cur_model, model_name) {
  imap_dfr(species_list, function(sp_label, sp) {
    plotYieldVsF(cur_model, species = sp, F_range = f_ranges[[sp]])$data %>%
      mutate(Species = sp_label, Model = model_name)
  })
})

plot_data$Species <- factor(plot_data$Species, levels = unname(species_list))

yields <- ggplot(plot_data, aes(x = F, y = yield, group = Model)) +
  geom_line(linewidth = 1, colour = "grey30") +
  geom_hline(yintercept = -Inf, linewidth = 0.5) +
  geom_vline(xintercept = -Inf, linewidth = 0.5) +
  facet_wrap(~ Species, scales = "free_y", ncol = 2, labeller = as_labeller(function(x) gsub("_", " ", x))) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 3),
                     expand = expansion(mult = c(0.05, 0.05))) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.02))) +
  labs(x = "Fishing mortality (1/yr)",
       y = "Yield (g/m²/yr)") +
  theme_classic() +
  theme(
    text = element_text(size = 17, family = "gillsans"),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 12, hjust = 0),
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 14),
    axis.title = element_text(size = 14),
    axis.line = element_blank(),
    legend.position = "none"
  )

# ggsave("yields.tiff", yields, width = 7, height = 6, dpi = 600, bg = "white")

# extracting information for Fmsy plots
peak_values <- plot_data %>%
  group_by(Species, Model) %>%
  slice_max(yield, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(Species, Model, F_peak = F, yield_peak = yield) %>%
  arrange(Species, Model)

peak_values
```

# Fmsy

```{r}
# Create the dataframe
fmsy_df <- data.frame(
  Species = c("bream", "perch", "pikeperch", "roach", "white_bream", 
              "small_fish", "ruffe"),
  mean = c(0.144, 0.122, 0.141, 0.204, 0.136, NA, NA),
  low = c(0.102, 0.086, 0.097, 0.145, 0.093, NA, NA),
  up = c(0.203, 0.177, 0.197, 0.284, 0.200, NA, NA)
)

# 1. Summarise model F_MSY (= F_peak) across the 5 models per species
fmsy_model_ci <- peak_values %>%
  group_by(Species) %>%
  summarise(
    F_MSY_model = mean(F_peak),
    model_low   = min(F_peak),          # or quantile(F_peak, 0.025)
    model_up    = max(F_peak),          # or quantile(F_peak, 0.975)
    .groups = "drop"
  )

cur_model <- models$M5
# 2. Lookup w_max per species
species_lookup <- data.frame(
  Species       = c("small_fish", "ruffe", "bream", "white_bream", "roach", "perch", "pikeperch"),
  w_max         = cur_model@species_params$w_max
)

# 3. Combine literature + model + w_max
fmsy_all <- fmsy_df %>%                      # literature: Species, mean, low, up
  left_join(fmsy_model_ci, by = "Species") %>%
  left_join(species_lookup, by = "Species")

print(fmsy_all)

# 4. Plot: w_max on x, both sources with their error bars

off <- 1.04  # multiplicative nudge; tune to taste

fmsy <- ggplot(fmsy_all, aes(x = w_max)) +
  # literature: shifted left
  geom_errorbar(aes(x = w_max / off, ymin = low, ymax = up),
                width = 100, color = "black", alpha = 0.5) +
  geom_point(aes(x = w_max / off, y = mean, color = Species,
                 shape = "Single species assessment"),
             size = 6) +
  # model: shifted right
  geom_errorbar(aes(x = w_max * off, ymin = model_low, ymax = model_up),
                width = 100, color = "grey40", alpha = 0.5) +
  geom_point(aes(x = w_max * off, y = F_MSY_model, color = Species,
                 shape = "Size spectrum model"),
             size = 6) +
  scale_y_continuous(limits = c(0, NA), breaks = seq(0, 0.6, 0.1)) +
  scale_color_manual(values = c("small_fish" = "#FCC300",
                                "ruffe" = "#44AA99", 
                                "white_bream" = "#661100",
                                "roach" = "#AA4499",
                                "perch" = "#2166ac",
                                "bream" = "#b2182b",
                                "pikeperch" = "#93AA00"),
                     breaks = c("small_fish", "ruffe", "bream", "white_bream",
                                "roach", "perch", "pikeperch"),
                     labels = function(x) gsub("_", " ", x)) +
  scale_shape_manual(values = c("Single species assessment" = 15,
                                "Size spectrum model" = 16),
                     breaks = c("Single species assessment", "Size spectrum model"),
                     labels = c("single species assessment", "size spectrum model")) +
  labs(x = expression(W[max]~"(g)"),
       y = expression(F[MSY]~"(year"^-1*")"),
       color = "Species",
       shape = "Source") +
  theme_classic() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

# ggsave("fmsy.tiff", fmsy, width = 7, height = 4, dpi = 600, bg = "white")
```

# Biomass

```{r}
# w_max per species (same across models; take from any one) 
species_wmax <- data.frame(
  Species = models$M5@species_params$species,
  w_max   = models$M5@species_params$w_max
)

# Observed biomass with upper/lower bounds (one set, from any model)
obs_biomass <- tibble(
  Species  = models$M5@species_params$species,
  obs      = models$M5@species_params$biomass_observed,
  obs_low  = models$M5@species_params$biomass_observed_low,
  obs_up   = models$M5@species_params$biomass_observed_up
) %>%
  filter(!is.na(obs)) %>%
  left_join(species_wmax, by = "Species")

#  Modelled biomass across all 5 models
model_biomass <- imap_dfr(models, function(m, model_name) {
  b <- plotBiomassVsSpecies(m)$data
  b %>%
    filter(Type == "Model") %>%
    mutate(Model = model_name)
}) %>%
  group_by(Species) %>%
  summarise(model_min = min(Biomass),
            model_max = max(Biomass),
            model_med = median(Biomass),
            .groups = "drop") %>%
  left_join(species_wmax, by = "Species")

# Order species by w_max consistently across both layers
lvl <- species_wmax %>% arrange(w_max) %>% pull(Species)
obs_biomass   <- mutate(obs_biomass,   Species = factor(Species, levels = lvl))
model_biomass <- mutate(model_biomass, Species = factor(Species, levels = lvl))

biomass1 <- ggplot() +
  # Observed range (error bar) + point
  geom_errorbar(data = obs_biomass,
                aes(x = w_max, ymin = obs_low, ymax = obs_up, color = Species),
                width = 0, linewidth = 2, alpha = 0.5,
                position = position_nudge(x = -0.04)) +
  geom_point(data = obs_biomass,
             aes(x = w_max, y = obs, color = Species, shape = "Observation"),
             size = 4, position = position_nudge(x = -0.04)) +
  # Modelled range across 5 models (error bar) + median point
  geom_errorbar(data = model_biomass,
                aes(x = w_max, ymin = model_min, ymax = model_max, color = Species),
                width = 0, linewidth = 2, alpha = 0.5,
                position = position_nudge(x = 0.02)) +
  geom_point(data = model_biomass,
             aes(x = w_max, y = model_med, color = Species, shape = "Model"),
             size = 4, position = position_nudge(x = 0.02)) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = c("small_fish" = "#FCC300", "ruffe" = "#44AA99",
                                "white_bream" = "#661100", "roach" = "#AA4499",
                                "perch" = "#2166ac", "bream" = "#b2182b",
                                "pikeperch" = "#93AA00"),
                     breaks = c("small_fish", "ruffe", "bream", "white_bream",
                                "roach", "perch", "pikeperch"),
                     labels = function(x) gsub("_", " ", x)) +
  scale_shape_manual(values = c("Observation" = 15, "Model" = 16),
                     breaks = c("Observation", "Model"),
                     labels = c("observation", "model")) +
  labs(x = "Maximum size (g)", y = "Biomass (g/m²)",
       color = "Species", shape = "Type") +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.text.y = element_text(size = 14),
    axis.title = element_text(size = 14),
    legend.position = "right",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 14)
  )

# ggsave("biomass1.tiff", biomass1, width = 6, height = 4, dpi = 600, bg = "white")
```

# Spectrum and feeding level

```{r}
# species size limits per model
get_limits <- function(m) {
  sp <- species_params(m)
  tibble(Species = sp$species, w_min = sp$w_min, w_max = sp$w_max)
}

# Spectrum
spectrum_data <- imap_dfr(models, function(m, model_name) {
  df <- plotSpectra(m, power = 1, return_data = TRUE)
  if ("Spectra" %in% names(df)) df <- rename(df, Species = Spectra)
  df %>% mutate(Model = model_name)
}) %>%
  mutate(group = interaction(Species, Model))

# Feeding level
feeding_data <- imap_dfr(models, function(m, model_name) {
  lims <- get_limits(m)
  getFeedingLevel(m) %>%
    as.data.frame() %>%
    rownames_to_column("Species") %>%
    pivot_longer(-Species, names_to = "w", values_to = "value") %>%
    mutate(w = as.numeric(w), Model = model_name) %>%
    left_join(lims, by = "Species") %>%
    filter(w >= w_min, w <= w_max)
}) %>%
  mutate(group = interaction(Species, Model))


plot_b <- ggplot(feeding_data, aes(x = w, y = value, color = Species, group = group)) +
  geom_line(linewidth = 1, alpha = 0.6) +
  geom_hline(yintercept = 0.2, linetype = "dashed", color = "red", linewidth = 1) +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.5)) +
  scale_color_manual(values = c("small_fish" = "#FCC300", "ruffe" = "#44AA99",
                                "white_bream" = "#661100", "roach" = "#AA4499",
                                "perch" = "#2166ac", "bream" = "#b2182b",
                                "pikeperch" = "#93AA00"),
                     breaks = c("small_fish", "ruffe", "bream", "white_bream",
                                "roach", "perch", "pikeperch"),
                     labels = function(x) gsub("_", " ", x)) +
  labs(x = "Size (g)", y = "Feeding level", color = "Species") +
  theme_classic() +
  theme(legend.position = "right",
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 12),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 15))

plot_a <- ggplot(spectrum_data, aes(x = w, y = value, color = Species,
                                    group = group, linetype = Species)) +
  geom_line(linewidth = 1, alpha = 0.6) +
  scale_x_log10() + scale_y_log10() +
  scale_color_manual(values = c("Resource" = "grey50", "small_fish" = "#FCC300", "ruffe" = "#44AA99",
                                "white_bream" = "#661100", "roach" = "#AA4499",
                                "perch" = "#2166ac", "bream" = "#b2182b",
                                "pikeperch" = "#93AA00"),
                     breaks = c("Resource", "small_fish", "ruffe", "bream", "white_bream",
                                "roach", "perch", "pikeperch"),
                     labels = function(x) gsub("_", " ", sub("Resource", "resource", x))) +
  scale_linetype_manual(values = c("Resource" = "dashed", "small_fish" = "solid", "ruffe" = "solid",
                                   "white_bream" = "solid", "roach" = "solid",
                                   "perch" = "solid", "bream" = "solid",
                                   "pikeperch" = "solid"),
                        breaks = c("Resource", "small_fish", "ruffe", "bream", "white_bream",
                                   "roach", "perch", "pikeperch"),
                        labels = function(x) gsub("_", " ", sub("Resource", "resource", x))) +
  labs(x = "Size (g)", y = "Biomass density (g/m²)", color = "Species", linetype = "Species") +
  theme_classic() +
  theme(legend.position = "right",
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 14),
        axis.text = element_text(size = 14),
        axis.title = element_text(size = 15))

spectra <- plot_a + (plot_b + theme(legend.position = "none")) +
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(tag_levels = "a", tag_suffix = ")")

# ggsave("spectra.tiff", spectra, width = 10, height = 4, dpi = 600, bg = "white")
```

# Growth curves

Checking the available size and age data across species

```{r}
# Read and select necessary columns
length_at_age <- read.csv("size_at_age.csv") %>%
  select(species = spp_name_lt, age = age, length = total_length, weight = fish_weight, std_length, year) %>% 
  mutate(species = recode(species,
                          "Eserys" = "perch",
                          "Karsis" = "bream",
                          "Kuoja" = "roach",
                          "Starkis" = "pikeperch"))


# Fit a model and fill missing values for each species
length_at_age <- length_at_age %>%
  group_by(species) %>%
  group_modify(~ {
    # Fit linear model if there are no missing values in total_length
    model <- lm(length ~ std_length, data = .x)
    
    # Extract coefficients
    intercept <- coef(model)[1]
    slope <- coef(model)[2]
    
    # Fill missing total_length
    .x <- .x %>%
      mutate(length = ifelse(is.na(length),
                                   intercept + (std_length * slope),
                                   length))
    return(.x)
  }) %>%
  ungroup()
 

colSums(is.na(length_at_age))

# length_at_age <- length_at_age %>% filter(year >= 1980 & year <=2000)

# Let's check the plot
ggplot(data = length_at_age) + 
  geom_point(mapping = aes(x = age, y = length, color=species), position = "jitter", size = 2)

length_at_age <- length_at_age %>% filter(length <= 100)

# Let's check the plot now
ggplot(data = length_at_age) + 
  geom_point(mapping = aes(x = age, y = weight, color=species), position = "jitter", size = 2)

# Prepare observed data
obs_data <- length_at_age %>%
  filter(!is.na(weight), !is.na(age)) %>%
  rename(Species = species)  # ADD THIS LINE to match growth_data

# Get max observed age per species
max_age_data <- obs_data %>%
  group_by(Species) %>%  # Change to Species
  summarise(max_age = max(age, na.rm = TRUE))
```

Model growth curves vs. empirical data

```{r}
species_order <- c("bream", "white_bream", "roach", "perch", "pikeperch")

# Extract growth curves from all 5 models
growth_data <- imap_dfr(models, function(m, model_name) {
  plotGrowthCurves(m, return_data = TRUE) %>%   # cols: Age, value, Species
    mutate(Model = model_name)
}) %>%
  filter(Species %in% species_order) %>%
  mutate(Species = factor(Species, levels = species_order),
         group   = interaction(Species, Model))


# Pull w_mat / w_max / age_mat from each model's species_params
ref_data <- imap_dfr(models, function(m, model_name) {
  sp <- species_params(m)
  tibble(Species = sp$species,
         w_mat   = sp$w_mat,
         w_max   = sp$w_max,
         age_mat = if (!is.null(sp$age_mat)) sp$age_mat else NA_real_)
}) %>%
  filter(Species %in% species_order) %>%
  group_by(Species) %>%
  summarise(w_mat = mean(w_mat, na.rm = TRUE),
            w_max = mean(w_max, na.rm = TRUE),
            age_mat = mean(age_mat, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(Species = factor(Species, levels = species_order))

# Max age per species (one line per facet)
max_age_data <- max_age_data %>%
  filter(Species %in% species_order) %>%
  mutate(Species = factor(Species, levels = species_order))

# obs_data ordering
obs_data <- obs_data %>%
  filter(Species %in% species_order) %>%
  mutate(Species = factor(Species, levels = species_order))

# --- Plot ---
growth <- ggplot() +
  geom_point(data = obs_data,
             aes(x = age, y = weight, color = Species),
             alpha = 0.3, position = position_jitter(width = 0.3), size = 1.5) +
  geom_line(data = growth_data,
            aes(x = Age, y = value, color = Species, group = group),
            linewidth = 1, alpha = 0.6) +
  geom_hline(data = ref_data, aes(yintercept = w_mat),
             linetype = "dashed", color = "black", alpha = 0.5) +
  geom_hline(data = ref_data, aes(yintercept = w_max),
             linetype = "dotted", color = "black", alpha = 0.5) +
  geom_vline(data = ref_data, aes(xintercept = age_mat),
             linetype = "dashed", color = "black", alpha = 0.5) +
  geom_vline(data = max_age_data, aes(xintercept = max_age),
             linetype = "dotted", color = "red", alpha = 0.5) +
  geom_hline(yintercept = -Inf, linewidth = 0.5) +
  geom_vline(xintercept = -Inf, linewidth = 0.5) +
  scale_color_manual(values = c("white_bream" = "#661100",
                                "roach"       = "#AA4499",
                                "perch"       = "#2166ac",
                                "bream"       = "#b2182b",
                                "pikeperch"   = "#93AA00")) +
  labs(x = "Age (years)", y = "Fish weight (g)", color = "Species") +
  facet_wrap(~Species, scales = "free_y", ncol = 2, labeller = as_labeller(function(x) gsub("_", " ", x))) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", hjust = 0, size = 11),
    legend.position = "none",
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

growth

growth_data %>%
  group_by(Species, Age) %>%
  summarise(spread = max(value) - min(value), .groups = "drop") %>%
  group_by(Species) %>%
  summarise(max_spread = max(spread))

# ggsave("growth.tiff", growth, width = 7, height = 6, dpi = 600, bg = "white")
```

# Diet

```{r}
cur_model <- read_rds("M5_kappa7_lambda2.05extMort01_0.333.rds")

# Get the diet data
diet_plot <- plotDiet(cur_model, 
                      species = c("small_fish", "ruffe", "bream", "white_bream", 
                                  "roach", "perch", "pikeperch"))
diet_data <- diet_plot$data

# Create custom plot with species colors and fixed x-axis
diet <- ggplot(diet_data, aes(x = size, y = Proportion, fill = Prey)) +
  geom_area() +
  facet_wrap(~ Predator, ncol = 2, labeller = as_labeller(function(x) gsub("_", " ", x))) +
  scale_x_log10() +
  scale_fill_manual(
  values = c("small_fish" = "#FCC300",
             "ruffe" = "#44AA99",
             "white_bream" = "#661100",
             "roach" = "#AA4499",
             "perch" = "#2166ac",
             "bream" = "#b2182b",
             "pikeperch" = "#93AA00",
             "Resource" = "#DDCC77",
             "External" = "grey30"),
  breaks = c("Resource", "small_fish", "ruffe", "bream", "white_bream", "roach", "perch", "pikeperch","External"),  # legend order
  labels = c("resource", "small fish", "ruffe", "bream", "white bream", "roach", "perch", "pikeperch","External")
) +
  labs(x = "Predator size (g)",
       y = "Proportion of diet",
       fill = "Prey") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", hjust = 0, size = 11),
    legend.position = "right",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

# ggsave("diet1.tiff", diet, width = 7, height = 6, dpi = 600, bg = "white")
```

# Mortality

```{r}
# Get death data
death_plot <- plotDeath(cur_model, 
                        species = c("small_fish", "ruffe", "bream", "white_bream", 
                                    "roach", "perch", "pikeperch"), proportion = F)
death_data <- death_plot$data

death_data$Prey <- factor(death_data$Prey,
                          levels = c("small_fish", "ruffe", "bream", "white_bream",
                                     "roach", "perch", "pikeperch"))
# Create plot with stacked areas
death <- ggplot(death_data, aes(x = w, y = value, fill = Legend)) +
  geom_area(position = "stack") +
  facet_wrap(~ Prey, ncol = 2, labeller = as_labeller(function(x) gsub("_", " ", x))) +
  scale_x_log10() +
  scale_fill_manual(values = c("small_fish" = "#FCC300",
                                "ruffe" = "#44AA99", 
                                "white_bream" = "#661100",
                                "roach" = "#AA4499",
                                "perch" = "#2166ac",
                                "bream" = "#b2182b",
                                "pikeperch" = "#93AA00",
                                "Resource" = "#DDCC77",
                                "External" = "grey",
                                "Fishing" = "red"),
  breaks = c("Resource", "small_fish", "ruffe", "bream", "white_bream", "roach", "perch", "pikeperch", "External", "Fishing"),  # legend order
  labels = c("resource", "small fish", "ruffe", "bream", "white bream", "roach", "perch", "pikeperch", "external", "fishing")
) +
  labs(x = "Size (g)",
       y = "Mortality rate (year⁻¹)",
       fill = "Cause") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", hjust = 0, size = 11),
    legend.position = "right",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

# ggsave("death1.tiff", death, width = 7, height = 6, dpi = 600, bg = "white")
```

# Biomass dynamics under three fishing effort scenarios

species sensitivity

```{r}
# effort levels
initial_effort <- c(0, 0, 0.111, 0.006, 0.131, 0.077, 0.083)

effort_multipliers <- list(
  "0.0"            = 0,
  "initial effort" = 1,
  "triple effort"  = 3
)

t_max <- 100

# Project every model at every effort level, extract biomass
get_biomass_df <- function(params, effort_mult, effort_label, model_name) {
  sim <- project(params, effort = initial_effort * effort_mult, t_max = t_max)

  getBiomass(sim) %>%
    as.data.frame() %>%
    rownames_to_column("Year") %>%
    mutate(Year = as.numeric(Year)) %>%
    pivot_longer(-Year, names_to = "Species", values_to = "Biomass") %>%
    mutate(Effort = effort_label, Model = model_name)
}

all_data <- imap_dfr(models, function(params, model_name) {
  imap_dfr(effort_multipliers, function(mult, label) {
    get_biomass_df(params, mult, label, model_name)
  })
})

# Summarize across the 5 models: median + min-max
summary_data <- all_data %>%
  group_by(Effort, Species, Year) %>%
  summarise(
    Biomass_med = median(Biomass),
    Biomass_min = min(Biomass),
    Biomass_max = max(Biomass),
    .groups = "drop"
  ) %>%
  mutate(Effort = factor(Effort,
                         levels = c("0.0", "initial effort", "triple effort")))

# Plot: 3 panels by effort, ribbon = min-max across models
species_colors <- c(
  "small_fish" = "#FCC300",
  "ruffe" = "#44AA99", 
  "bream" = "#b2182b",
  "white_bream" = "#661100",
  "roach" = "#AA4499",
  "perch" = "#2166ac",
  "pikeperch" = "#93AA00"
)

species_order <- c("small_fish", "ruffe", "bream", "white_bream",
                   "roach", "perch", "pikeperch")

summary_data$Species <- factor(summary_data$Species, levels = species_order)

biomass_time <- ggplot(summary_data, aes(x = Year, color = Species, fill = Species)) +
  geom_ribbon(aes(ymin = Biomass_min, ymax = Biomass_max),
              alpha = 0.2, color = NA) +
  geom_line(aes(y = Biomass_med), linewidth = 1) +
  facet_wrap(~ Effort, ncol = 1, scales = "free_y",
             labeller = labeller(Effort = c("0.0" = "a) F reduced to 0",
                                            "initial effort" = "b) F stable",
                                            "triple effort" = "c) F increased 3 times"))) +
  coord_cartesian(xlim = c(0, 100)) +
  scale_color_manual(values = species_colors,
                   labels = c(small_fish = "small fish",
                              white_bream = "white bream")) +
  scale_fill_manual(values = species_colors,
                  labels = c(small_fish = "small fish",
                             white_bream = "white bream")) +
  labs(x = "Time (years)", y = "Biomass (g)",
       color = "Species", fill = "Species") +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 12, hjust = 0),
    legend.position = "right",
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 12),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    panel.border = element_blank(),
    axis.line.x = element_line(linewidth = 0.5, color = "black"),
    axis.line.y = element_line(linewidth = 0.5, color = "black")
  )

# ggsave("bio_timeS.tiff", biomass_time, width = 6, height = 6, dpi = 600)
```
