---
title: "w_mat change"
author: "Elyza Pilipaityte"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---


clear memory

```{r}
rm(list=ls())

```

Install mizer, load libraries

```{r}
# remotes::install_github("sizespectrum/mizerExperimental", ref = "tuneMR_w_max", force=T)
# remotes::install_github("sizespectrum/mizerExperimental")
# library(mizer)
# install.packages("ggh4x")
library(ggh4x)
library(mizerExperimental)
library(tidyverse)
library(assertthat)
library(reshape2)
library(patchwork)
```

```{r}
My_theme <- theme(
    panel.border = element_blank(),
    panel.grid = element_blank(),
    panel.background = element_blank(),        # removes grey panel background
    plot.background = element_blank(),         # removes grey plot background
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.line.x.bottom = element_line(color = "gray5", linewidth = 0.3),
    axis.line.y.left = element_line(color = "gray5", linewidth = 0.3),
    text = element_text(size = 15, family = "gillsans"))# font

initial_effort <- c(0, 0, 0.111, 0.006, 0.131, 0.077, 0.083)
```

# Simulation of decreasing size at maturity

```{r}
# List of all models for uncertainty
model_files <- c(
  "M6_kappa8_lambda2.05extMort01_0.333.rds",
  "M4_kappa7_lambda2.05extMort01_0.25.rds",
  "M5_kappa7_lambda2.05extMort01_0.333.rds",
  "M12_kappa7_lambda2.05extMort02_0.333.rds",
  "M7_kappa9_lambda2.05extMort01_0.333.rds"
)

# w_mat decrease scenarios
scenarios <- list(
  all_species = c(0.01, 0.01, 0.01, 0.01, 0.01, 0.01, 0.01),
  species_3_4_5_6_7 = c(0, 0, 0.01, 0.01, 0.01, 0.01, 0.01),
  species_7_only = c(0, 0, 0, 0, 0, 0, 0.01),
  species_6_only = c(0, 0, 0, 0, 0, 0.01, 0),
  species_5_only = c(0, 0, 0, 0, 0.01, 0, 0),
  species_4_only = c(0, 0, 0, 0.01, 0, 0, 0),
  species_3_only = c(0, 0, 0.01, 0, 0, 0, 0),
  species_2_only = c(0, 0.01, 0, 0, 0, 0, 0),
  species_1_only = c(0.01, 0, 0, 0, 0, 0, 0),
  no_change = c(0, 0, 0, 0, 0, 0, 0)
)

# Functions 
psi_at_t <- function(params, t) {
  op <- params@other_params$psi
  list(psi = op[t + 1, , ])
}

timeVaryingERepro <- function(params, n, n_pp, n_other, t, e, ...) {
  e[e < 0] <- 0
  pars <- psi_at_t(params, t)
  psi <- pars$psi
  psi * e 
}

# Store all results
all_model_results <- list()

# Loop through each model
for (model_file in model_files) {
  cat("\n========================================\n")
  cat("Processing model:", model_file, "\n")
  cat("========================================\n")
  
  # Load the model
  cur_model <- read_rds(model_file)
  
  # Initial projection to get starting conditions
  sim <- project(cur_model, t_max = 300, dt = 0.2, effort = initial_effort)
  initial_n <- sim@n[300,,]
  initial_n_pp <- sim@n_pp[300,]
  
  # Store results for this model
  model_results <- list()
  
  # Looping through scenarios for one model at a time
  for (scenario_name in names(scenarios)) {
    cat("  Running scenario:", scenario_name, "\n")
    
    decrease_rate <- scenarios[[scenario_name]]
    model_scenario <- cur_model
    
    # Creating psi array
    times <- 0:100
    species <- species_params(model_scenario)$species
    n_sp <- length(species)
    w <- model_scenario@w
    
    psi_array <- array(NA, dim = c(length(times), length(species), length(w)),
                       dimnames = list(time = times, sp = species, w = w))
    w_mat_array <- array(NA, dim = c(length(times), length(species)),
                         dimnames = list(time = times, sp = species))
    w_mat25_array <- w_mat_array
    w_max_array <- w_mat_array
    
    initial_w_mat   <- model_scenario@species_params$w_mat
    initial_w_mat25 <- model_scenario@species_params$w_mat25
    initial_w_max   <- model_scenario@species_params$w_max
    
    w_mat_current   <- initial_w_mat
    w_mat25_current <- initial_w_mat25
    w_max_current   <- initial_w_max
    
    for (i in seq_along(times)) {
      w_mat_array[i, ]   <- w_mat_current
      w_mat25_array[i, ] <- w_mat25_current
      w_max_array[i, ]   <- w_max_current
      w_mat_current      <- w_mat_current * (1 - decrease_rate)
      w_mat25_current    <- w_mat25_current * (1 - decrease_rate)
      w_max_current      <- w_max_current * (1 - decrease_rate)
    }
    
    m <- model_scenario@species_params$m
    n <- model_scenario@species_params$n
    
    for (t in seq_along(times)) {
      for (sp in seq_len(n_sp)) {
        w_mat_t   <- w_mat_array[t, sp]
        w_mat25_t <- w_mat25_array[t, sp]
        w_repro_max_sp <- w_max_array[t, sp]
        
        if (!is.na(w_mat_t) && !is.na(w_mat25_t) && w_mat_t > w_mat25_t) {
          U        <- log(3) / log(w_mat_t / w_mat25_t)
          maturity <- (1 + (w / w_mat_t)^(-U))^(-1)
          maturity[maturity < 1e-08] <- 0 
          
          repro_prop <- (w / w_repro_max_sp)^(m[sp] - n[sp])
          repro_prop[repro_prop > 1] <- 1
          
          psi <- maturity * repro_prop
          psi[psi > 1] <- 1
          psi[w > w_repro_max_sp] <- 1
          
          psi_array[t, sp, ] <- psi
        } else {
          psi_array[t, sp, ] <- 0
        }
      }
    }
    
    model_scenario@other_params$psi <- psi_array
    model_scenario <- setRateFunction(model_scenario, "ERepro", "timeVaryingERepro")
    
    # Run simulation
    sim <- project(model_scenario, t_max = 50, effort = initial_effort, dt = 0.1, 
                   initial_n = initial_n, initial_n_pp = initial_n_pp)
    
    model_results[[scenario_name]] <- sim
  }
  
  # Storing results
  all_model_results[[model_file]] <- model_results
}

cat("\nAll simulations complete!\n")

```

upper and lower values used
```{r}
# Extracting biomass data from all models
combined_biomass_data <- data.frame()

for (model_name in names(all_model_results)) {
  for (scenario_name in names(all_model_results[[model_name]])) {
    sim <- all_model_results[[model_name]][[scenario_name]]
    biomass <- getBiomass(sim)
    
    biomass_long <- as.data.frame(biomass) %>%
      mutate(time = as.numeric(rownames(biomass))) %>%
      pivot_longer(cols = -time, names_to = "species", values_to = "biomass") %>%
      mutate(
        model = model_name,
        scenario = scenario_name
      )
    
    combined_biomass_data <- bind_rows(combined_biomass_data, biomass_long)
  }
}

# Summary statistics with min and max values
biomass_summary <- combined_biomass_data %>%
  group_by(scenario, species, time) %>%
  summarise(
    mean_biomass = mean(biomass),
    median_biomass = median(biomass),
    sd_biomass = sd(biomass),
    min_biomass = min(biomass),      # Minimum across models
    max_biomass = max(biomass),      # Maximum across models
    .groups = 'drop'
  )

# Calculating percentage change with min/max bounds
biomass_change_summary <- combined_biomass_data %>%
  group_by(model, scenario, species) %>%
  mutate(biomass_change_percent = (biomass - first(biomass)) / first(biomass) * 100) %>%
  ungroup() %>%
  group_by(scenario, species, time) %>%
  summarise(
    mean_change = mean(biomass_change_percent),
    median_change = median(biomass_change_percent),
    min_change = min(biomass_change_percent),      # minimum value across models
    max_change = max(biomass_change_percent),      # maximum value across models
    .groups = 'drop'
  )
 
# Better scenario labels
biomass_change_summary <- biomass_change_summary %>%
  filter(scenario != "no_change") %>%
   mutate(scenario_label = case_when(
    scenario == "no_change" ~ "No decline in w_mat",
    scenario == "all_species" ~ "all 7 spp.",
    scenario == "species_3_4_5_6_7" ~ "5 commercial spp.",
    scenario == "species_1_only" ~ "small fish",
    scenario == "species_2_only" ~ "ruffe",
    scenario == "species_3_only" ~ "bream",
    scenario == "species_4_only" ~ "white bream",
    scenario == "species_5_only" ~ "roach",
    scenario == "species_6_only" ~ "perch",
    scenario == "species_7_only" ~ "pikeperch",
    TRUE ~ scenario
  )) %>%
  mutate(scenario_label = factor(scenario_label, levels = c(
    "all 7 spp.",
    "5 commercial spp.",
    "pikeperch",
    "perch",
    "roach",
    "bream",
    "white bream",
    "ruffe",
    "small fish"
  )))

# Species colors
species_colors <- c(
  "small_fish" = "#FCC300",
  "ruffe" = "#44AA99", 
  "bream" = "#b2182b",
  "white_bream" = "#661100",
  "roach" = "#AA4499",
  "perch" = "#2166ac",
  "pikeperch" = "#93AA00"
)

# plot 
p0 <- ggplot(biomass_change_summary, aes(x = time, y = mean_change, color = species, fill = species)) +
  geom_ribbon(aes(ymin = min_change, ymax = max_change),
              alpha = 0.2, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  facet_wrap2(~ scenario_label, scales = "fixed", ncol = 1, 
            axes = "all", remove_labels = "x") +
  labs(x = "Time (years)", 
       y = "Biomass change (%)", 
       color = "Species",
       fill = "Species") +
  scale_color_manual(values = species_colors) +
  scale_fill_manual(values = species_colors) +
  theme_minimal() +
theme(
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),

    axis.line.x.bottom = element_line(color = "black", linewidth = 0.5),
    axis.line.y.left = element_line(color = "black", linewidth = 0.5),
    strip.background = element_blank(),
    strip.text = element_text(size = 10, face = "bold", hjust = 0, margin = margin(b = 5)),
    
    # Axis styling
    axis.line = element_line(color = "black", linewidth = 0.5),
    axis.text = element_text(size = 9, color = "black"),
    axis.title = element_text(size = 11, face = "bold"),
    axis.ticks = element_line(color = "black", linewidth = 0.5),
    
    # Legend
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    legend.key = element_blank(),
    legend.background = element_blank(),
    
    # Overall plot
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(10, 10, 10, 10),
    plot.subtitle = element_text(size = 9, hjust = 0.5)
  )

```

# biomass matrix

```{r}
# Calculate relative biomass change for all models and scenarios
calculate_change_all_models <- function(all_model_results, time_point = 50) {
  
  # Initialize list to store results from each model
  model_changes <- list()
  
  for (model_name in names(all_model_results)) {
    # Extract results for this model
    model_results <- all_model_results[[model_name]]
    
    # Calculate changes for each scenario
    all_decrease <- (getBiomass(model_results$all_species)[time_point, ] - 
                     getBiomass(model_results$all_species)[1, ]) / 
                     getBiomass(model_results$all_species)[1, ]
    
    five_species_decrease <- (getBiomass(model_results$species_3_4_5_6_7)[time_point, ] - 
                              getBiomass(model_results$species_3_4_5_6_7)[1, ]) / 
                              getBiomass(model_results$species_3_4_5_6_7)[1, ]
    
    pikeperch_decrease <- (getBiomass(model_results$species_7_only)[time_point, ] - 
                           getBiomass(model_results$species_7_only)[1, ]) / 
                           getBiomass(model_results$species_7_only)[1, ]
    
    perch_decrease <- (getBiomass(model_results$species_6_only)[time_point, ] - 
                       getBiomass(model_results$species_6_only)[1, ]) / 
                       getBiomass(model_results$species_6_only)[1, ]
    
    roach_decrease <- (getBiomass(model_results$species_5_only)[time_point, ] - 
                       getBiomass(model_results$species_5_only)[1, ]) / 
                       getBiomass(model_results$species_5_only)[1, ]
    
    wbream_decrease <- (getBiomass(model_results$species_4_only)[time_point, ] - 
                        getBiomass(model_results$species_4_only)[1, ]) / 
                        getBiomass(model_results$species_4_only)[1, ]
    
    bream_decrease <- (getBiomass(model_results$species_3_only)[time_point, ] - 
                       getBiomass(model_results$species_3_only)[1, ]) / 
                       getBiomass(model_results$species_3_only)[1, ]
    
    ruffe_decrease <- (getBiomass(model_results$species_2_only)[time_point, ] - 
                       getBiomass(model_results$species_2_only)[1, ]) / 
                       getBiomass(model_results$species_2_only)[1, ]
    
    small_fish_decrease <- (getBiomass(model_results$species_1_only)[time_point, ] - 
                            getBiomass(model_results$species_1_only)[1, ]) / 
                            getBiomass(model_results$species_1_only)[1, ]
    
    no_change <- (getBiomass(model_results$no_change)[time_point, ] - 
                  getBiomass(model_results$no_change)[1, ]) / 
                  getBiomass(model_results$no_change)[1, ]
    
    # Combine into matrix
    model_change_matrix <- rbind(
      small_fish_decrease, 
      ruffe_decrease, 
      bream_decrease, 
      wbream_decrease, 
      roach_decrease, 
      perch_decrease, 
      pikeperch_decrease, 
      five_species_decrease, 
      all_decrease, 
      no_change
    )
    
    # Store with model name
    model_changes[[model_name]] <- model_change_matrix
  }
  
  return(model_changes)
}

# Calculate changes for all models
all_model_changes <- calculate_change_all_models(all_model_results, time_point = 50)

# Convert to long format with all models
changes_long_all <- data.frame()

for (model_name in names(all_model_changes)) {
  change_matrix <- all_model_changes[[model_name]] * 100  # Convert to percent
  change_long <- melt(change_matrix)
  colnames(change_long) <- c("Scenario", "Species", "Change_percent")
  change_long$Model <- model_name
  changes_long_all <- rbind(changes_long_all, change_long)
}

# Calculate summary statistics with MIN/MAX
changes_summary <- changes_long_all %>%
  group_by(Scenario, Species) %>%
  summarise(
    mean_change = mean(Change_percent),
    median_change = median(Change_percent),
    min_change = min(Change_percent),      # Minimum across models
    max_change = max(Change_percent),      # Maximum across models
    range_width = max(Change_percent) - min(Change_percent),  # Width of range
    .groups = 'drop'
  )

# Better scenario labels
changes_summary <- changes_summary %>%
  mutate(Scenario_label = case_when(
    Scenario == "small_fish_decrease" ~ "small fish",
    Scenario == "ruffe_decrease" ~ "ruffe",
    Scenario == "bream_decrease" ~ "bream",
    Scenario == "wbream_decrease" ~ "white bream",
    Scenario == "roach_decrease" ~ "roach",
    Scenario == "perch_decrease" ~ "perch",
    Scenario == "pikeperch_decrease" ~ "pikeperch",
    Scenario == "five_species_decrease" ~ "5 commercial spp.",
    Scenario == "all_decrease" ~ "all 7 spp.",
    Scenario == "no_change" ~ "no change",
    TRUE ~ as.character(Scenario)
  ))

# Reorder scenarios
changes_summary$Scenario_label <- factor(changes_summary$Scenario_label,
                                          levels = c("no change", "small fish", "ruffe", 
                                                     "bream", "white bream", "roach", 
                                                     "perch", "pikeperch", "5 commercial spp.", 
                                                     "all 7 spp."))
```

```{r}
# After running fishing scenarios, save the summary
changes_summary_fishing <- changes_summary %>%
  filter(Scenario_label != "no change") %>%
  mutate(
    Fishing = "With fishing",
    label_text = case_when(
      round(min_change, 0) == round(max_change, 0) ~ sprintf("%.0f", min_change),
      TRUE ~ sprintf("%.0f – %.0f", min_change, max_change)
    )
  )
```

```{r}
# Filter out "No change" scenario and create custom label
changes_summary <- changes_summary_fishing %>%
  filter(Scenario_label != "no change") %>%
  mutate(
    min_r = round(min_change, 0),
    max_r = round(max_change, 0),
    min_r = ifelse(min_r == 0, 0, min_r),
    max_r = ifelse(max_r == 0, 0, max_r),
    label_text = case_when(
      min_r == max_r ~ sprintf("%.0f", min_r),
      TRUE ~ sprintf("%.0f – %.0f", min_r, max_r)
    )
  ) %>%
  select(-min_r, -max_r)

# Reorder scenarios (without "No change")
changes_summary$Scenario_label <- factor(changes_summary$Scenario_label,
                                          levels = c("small fish", "ruffe", 
                                                     "bream", "white bream", "roach", 
                                                     "perch", "pikeperch", "5 commercial spp.", 
                                                     "all 7 spp."))

# Create plot
p_fisihng <- ggplot(changes_summary, aes(x = Species, y = Scenario_label, fill = mean_change)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = label_text), size = 3.5, lineheight = 1.1) +  # Increased size slightly
  scale_fill_gradient2(low = "#56b4e9", mid = "white", high = "#E69F00", 
                       midpoint = 0, name = "Mean\nChange (%)") +
  geom_hline(yintercept = 7.5, linewidth = 1.5, color = "grey30") +
  labs(x = "Species", 
       y = "w_mat Decrease Scenario",
       title = "Biomass Change at Year 50 (%)",
       subtitle = "Range (min – max) across 5 model parameterizations") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 9, face = "bold"),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA)
  )

```
I ran all the code for changes in diets, mortality, and size spectrum first, because I did not create separate code chunks for fished and unfished scenarios. Sometimes I rename the plot and save it so I can combine both the fished and unfished plots.

# Without fishing

Unfished biomass in five models
I want to check unfished biomass in equilibrium conditions
```{r}
eq_biomass_all <- list()

for (model_file in model_files) {
  cur_model <- read_rds(model_file)

  # with fishing
  sim_with_fishing <- project(cur_model, t_max = 300, effort = initial_effort, dt = 0.2)
  initial_n_fishing    <- sim_with_fishing@n[300,,]
  initial_n_pp_fishing <- sim_with_fishing@n_pp[300,]

  # no fishing to steady state
  sim_no_fishing <- project(cur_model, t_max = 1000, effort = 0, dt = 0.1,
                            initial_n = initial_n_fishing,
                            initial_n_pp = initial_n_pp_fishing)

  # Biomass at the last time step
  bm <- getBiomass(sim_no_fishing)
  eq_biomass_all[[model_file]] <- bm[nrow(bm), ]
}

# Combine all 5 models into one table
eq_biomass_df <- do.call(rbind, eq_biomass_all)
print(eq_biomass_df)

cv_pct <- apply(eq_biomass_df, 2, function(x) sd(x)/mean(x)*100)
round(range(cv_pct), 0)   # gives the low–high CV across species

var_pct_min <- apply(eq_biomass_df, 2, function(x) (max(x) - min(x)) / min(x) * 100)
round(range(var_pct_min), 0)

fished_biomass_all <- list()

for (model_file in model_files) {
  cur_model <- read_rds(model_file)

  # 300-year run with fishing
  sim_with_fishing <- project(cur_model, t_max = 300, effort = initial_effort, dt = 0.2)

  # Biomass at the last time step
  bm <- getBiomass(sim_with_fishing)
  fished_biomass_all[[model_file]] <- bm[nrow(bm), ]
}

# Combine all 5 models into one table (rows = models, cols = species)
fished_biomass_df <- do.call(rbind, fished_biomass_all)
print(fished_biomass_df)

# Between-model spread per species
cv_pct <- apply(fished_biomass_df, 2, function(x) sd(x)/mean(x)*100)
round(range(cv_pct), 0)

var_pct_min <- apply(fished_biomass_df, 2, function(x) (max(x) - min(x)) / min(x) * 100)
round(range(var_pct_min), 0)
```

maturity size decrease simulation

```{r}
# List of all models to test
model_files <- c(
  "M6_kappa8_lambda2.05extMort01_0.333.rds",
  "M4_kappa7_lambda2.05extMort01_0.25.rds",
  "M5_kappa7_lambda2.05extMort01_0.333.rds",
  "M12_kappa7_lambda2.05extMort02_0.333.rds",
  "M7_kappa9_lambda2.05extMort01_0.333.rds"
)

# scenarios (same as before)
scenarios <- list(
  all_species = c(0.01, 0.01, 0.01, 0.01, 0.01, 0.01, 0.01),
  species_3_4_5_6_7 = c(0, 0, 0.01, 0.01, 0.01, 0.01, 0.01),
  species_7_only = c(0, 0, 0, 0, 0, 0, 0.01),
  species_6_only = c(0, 0, 0, 0, 0, 0.01, 0),
  species_5_only = c(0, 0, 0, 0, 0.01, 0, 0),
  species_4_only = c(0, 0, 0, 0.01, 0, 0, 0),
  species_3_only = c(0, 0, 0.01, 0, 0, 0, 0),
  species_2_only = c(0, 0.01, 0, 0, 0, 0, 0),
  species_1_only = c(0.01, 0, 0, 0, 0, 0, 0),
  no_change = c(0, 0, 0, 0, 0, 0, 0)
)

# Functions (same as before)
psi_at_t <- function(params, t) {
  op <- params@other_params$psi
  list(psi = op[t + 1, , ])
}

timeVaryingERepro <- function(params, n, n_pp, n_other, t, e, ...) {
  e[e < 0] <- 0
  pars <- psi_at_t(params, t)
  psi <- pars$psi
  psi * e 
}

# Store ALL results
all_model_results <- list()

# Loop through each model
for (model_file in model_files) {
  cat("\n========================================\n")
  cat("Processing model:", model_file, "\n")
  cat("========================================\n")
  
  # Load the model
  cur_model <- read_rds(model_file)
  
  # Step 1: Initial projection WITH fishing to establish baseline
  cat("  Step 1: Running with fishing to establish baseline...\n")
  sim_with_fishing <- project(cur_model, t_max = 300, effort = initial_effort, dt = 0.2)
  
  initial_n_fishing <- sim_with_fishing@n[300,,]
  initial_n_pp_fishing <- sim_with_fishing@n_pp[300,]
  
  # Step 2: Long projection WITHOUT fishing to reach steady state
  cat("  Step 2: Running without fishing to reach steady state (1000 years)...\n")
  sim_no_fishing <- project(cur_model, t_max = 1000, effort = 0, dt = 0.1, 
                            initial_n = initial_n_fishing, 
                            initial_n_pp = initial_n_pp_fishing)
  
  # Extract steady state conditions (end of no-fishing period)
  initial_n <- sim_no_fishing@n[1000,,]
  initial_n_pp <- sim_no_fishing@n_pp[1000,]
  
  cat("  Step 3: Steady state reached. Starting scenario simulations...\n")
  
  # Store results for this model
  model_results <- list()
  
  # Loop through scenarios for this model
  for (scenario_name in names(scenarios)) {
    cat("    Running scenario:", scenario_name, "\n")
    
    decrease_rate <- scenarios[[scenario_name]]
    model_scenario <- cur_model
    
    # Create psi array
    times <- 0:100
    species <- species_params(model_scenario)$species
    n_sp <- length(species)
    w <- model_scenario@w
    
    psi_array <- array(NA, dim = c(length(times), length(species), length(w)),
                       dimnames = list(time = times, sp = species, w = w))
    w_mat_array <- array(NA, dim = c(length(times), length(species)),
                         dimnames = list(time = times, sp = species))
    w_mat25_array <- w_mat_array
    w_max_array <- w_mat_array
    
    initial_w_mat   <- model_scenario@species_params$w_mat
    initial_w_mat25 <- model_scenario@species_params$w_mat25
    initial_w_max   <- model_scenario@species_params$w_max
    
    w_mat_current   <- initial_w_mat
    w_mat25_current <- initial_w_mat25
    w_max_current   <- initial_w_max
    
    # Build the declining w_mat arrays
    for (i in seq_along(times)) {
      w_mat_array[i, ]   <- w_mat_current
      w_mat25_array[i, ] <- w_mat25_current
      w_max_array[i, ]   <- w_max_current
      w_mat_current      <- w_mat_current * (1 - decrease_rate)
      w_mat25_current    <- w_mat25_current * (1 - decrease_rate)
      w_max_current      <- w_max_current * (1 - decrease_rate)
    }
    
    m <- model_scenario@species_params$m
    n <- model_scenario@species_params$n
    
    # Calculate psi for each time step
    for (t in seq_along(times)) {
      for (sp in seq_len(n_sp)) {
        w_mat_t   <- w_mat_array[t, sp]
        w_mat25_t <- w_mat25_array[t, sp]
        w_repro_max_sp <- w_max_array[t, sp]
        
        if (!is.na(w_mat_t) && !is.na(w_mat25_t) && w_mat_t > w_mat25_t) {
          U        <- log(3) / log(w_mat_t / w_mat25_t)
          maturity <- (1 + (w / w_mat_t)^(-U))^(-1)
          maturity[maturity < 1e-08] <- 0 
          
          repro_prop <- (w / w_repro_max_sp)^(m[sp] - n[sp])
          repro_prop[repro_prop > 1] <- 1
          
          psi <- maturity * repro_prop
          psi[psi > 1] <- 1
          psi[w > w_repro_max_sp] <- 1
          
          psi_array[t, sp, ] <- psi
        } else {
          psi_array[t, sp, ] <- 0
        }
      }
    }
    
    model_scenario@other_params$psi <- psi_array
    model_scenario <- setRateFunction(model_scenario, "ERepro", "timeVaryingERepro")
    
    # Run simulation with NO FISHING, starting from steady state
    sim <- project(model_scenario, t_max = 50, effort = 0, dt = 0.1,
                   initial_n = initial_n, initial_n_pp = initial_n_pp)
    
    model_results[[scenario_name]] <- sim
  }
  
  # Store this model's results
  all_model_results[[model_file]] <- model_results
  
  cat("  Completed model:", model_file, "\n")
}

cat("\n========================================\n")
cat("All simulations complete!\n")
cat("========================================\n")
```

upper and lower values used biomass change through time
```{r}
species_order <- c("small_fish", "ruffe", "bream", "white_bream", 
                   "roach", "perch", "pikeperch")

# Extract biomass data from all models
combined_biomass_data <- data.frame()

for (model_name in names(all_model_results)) {
  for (scenario_name in names(all_model_results[[model_name]])) {
    sim <- all_model_results[[model_name]][[scenario_name]]
    biomass <- getBiomass(sim)
    
    biomass_long <- as.data.frame(biomass) %>%
      mutate(time = as.numeric(rownames(biomass))) %>%
      pivot_longer(cols = -time, names_to = "species", values_to = "biomass") %>%
      mutate(
        model = model_name,
        scenario = scenario_name
      )
    
    combined_biomass_data <- bind_rows(combined_biomass_data, biomass_long)
  }
}

# Calculate summary statistics with MIN/MAX instead of CI
biomass_summary <- combined_biomass_data %>%
  group_by(scenario, species, time) %>%
  summarise(
    mean_biomass = mean(biomass),
    median_biomass = median(biomass),
    sd_biomass = sd(biomass),
    min_biomass = min(biomass),      # Minimum across models
    max_biomass = max(biomass),      # Maximum across models
    .groups = 'drop'
  )

# Calculate percentage change with MIN/MAX bounds
biomass_change_summary <- combined_biomass_data %>%
  group_by(model, scenario, species) %>%
  mutate(biomass_change_percent = (biomass - first(biomass)) / first(biomass) * 100) %>%
  ungroup() %>%
  group_by(scenario, species, time) %>%
  summarise(
    mean_change = mean(biomass_change_percent),
    median_change = median(biomass_change_percent),
    min_change = min(biomass_change_percent),      # Minimum value across models
    max_change = max(biomass_change_percent),      # Maximum value across models
    .groups = 'drop'
  )
 
# Better scenario labels
biomass_change_summary <- biomass_change_summary %>%
  mutate(scenario_label = case_when(
    scenario == "no_change" ~ "No decline in w_mat",
    scenario == "all_species" ~ "all 7 spp.",
    scenario == "species_3_4_5_6_7" ~ "5 commercial spp.",
    scenario == "species_1_only" ~ "small fish",
    scenario == "species_2_only" ~ "ruffe",
    scenario == "species_3_only" ~ "bream",
    scenario == "species_4_only" ~ "white bream",
    scenario == "species_5_only" ~ "roach",
    scenario == "species_6_only" ~ "perch",
    scenario == "species_7_only" ~ "pikeperch",
    TRUE ~ scenario
  )) %>%
  mutate(scenario_label = factor(scenario_label, levels = c(
    "all 7 spp.",
    "5 commercial spp.",
    "pikeperch",
    "perch",
    "roach",
    "bream",
    "white bream",
    "ruffe",
    "small fish"
  )))

# Create the plot with no fishing 
p_noF <- biomass_change_summary %>%
  filter(scenario != "no_change") %>%
  ggplot(aes(x = time, y = mean_change, color = species, fill = species)) +
  geom_ribbon(aes(ymin = min_change, ymax = max_change),
              alpha = 0.2, color = NA) +
  coord_cartesian(xlim = c(0, 50)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  facet_wrap2(~ scenario_label, scales = "fixed", ncol = 1, 
            axes = "all", remove_labels = "x") +
  labs(x = "Time (years)", 
       y = "Biomass change (%)", 
       color = "Species",
       fill = "Species") +
scale_color_manual(values = species_colors, breaks = species_order,
                     labels = function(x) gsub("_", " ", x)) +
  scale_fill_manual(values = species_colors, breaks = species_order,
                    labels = function(x) gsub("_", " ", x)) +
  theme_minimal() +
theme(
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_blank(),

    axis.line.x.bottom = element_line(color = "black", linewidth = 0.5),
    axis.line.y.left = element_line(color = "black", linewidth = 0.5),
    
    # Strip (facet labels) - minimal styling
    strip.background = element_blank(),
    strip.text = element_text(size = 10, face = "bold", hjust = 0, margin = margin(b = 5)),
    
    # Axis styling
    axis.line = element_line(color = "black", linewidth = 0.5),
    axis.text = element_text(size = 9, color = "black"),
    axis.title = element_text(size = 11, face = "bold"),
    axis.ticks = element_line(color = "black", linewidth = 0.5),
    
    # Legend
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    legend.key = element_blank(),
    legend.background = element_blank(),
    
    # Overall plot
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin = margin(10, 10, 10, 10),
    plot.subtitle = element_text(size = 9, hjust = 0.5)
  )

```

combined plot of biomass change through time

```{r}
((p_noF + theme(legend.position = "none")) + 
 (p0 + theme(axis.title.y = element_blank()))) +
  plot_layout(guides = "collect")

combined <- ((p0 + 
                theme(legend.position = "none", 
                      axis.title.x = element_blank(),
                      plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) +
                ggtitle("With fishing")) + 
             (p_noF + 
                theme(axis.title.y = element_blank(), 
                      axis.title.x = element_blank(),
                      plot.title = element_text(hjust = 0.5, size = 12, face = "bold")) +
                ggtitle("Without fishing"))) +
  plot_layout(guides = "collect") +
  plot_annotation(
    caption = "Time (years)",
    theme = theme(plot.caption = element_text(hjust = 0.45, size = 11, 
                                               face = "bold", color = "black"))
  )

# ggsave("biomass_through_time.tiff", combined, width = 7.5, height = 10, dpi = 600, bg = "white")
```

# biomass matrix

```{r}
# Calculate relative biomass change for all models and scenarios
calculate_change_all_models <- function(all_model_results, time_point = 50) {
  
  # Initialize list to store results from each model
  model_changes <- list()
  
  for (model_name in names(all_model_results)) {
    # Extract results for this model
    model_results <- all_model_results[[model_name]]
    
    # Calculate changes for each scenario
    all_decrease <- (getBiomass(model_results$all_species)[time_point, ] - 
                     getBiomass(model_results$all_species)[1, ]) / 
                     getBiomass(model_results$all_species)[1, ]
    
    five_species_decrease <- (getBiomass(model_results$species_3_4_5_6_7)[time_point, ] - 
                              getBiomass(model_results$species_3_4_5_6_7)[1, ]) / 
                              getBiomass(model_results$species_3_4_5_6_7)[1, ]
    
    pikeperch_decrease <- (getBiomass(model_results$species_7_only)[time_point, ] - 
                           getBiomass(model_results$species_7_only)[1, ]) / 
                           getBiomass(model_results$species_7_only)[1, ]
    
    perch_decrease <- (getBiomass(model_results$species_6_only)[time_point, ] - 
                       getBiomass(model_results$species_6_only)[1, ]) / 
                       getBiomass(model_results$species_6_only)[1, ]
    
    roach_decrease <- (getBiomass(model_results$species_5_only)[time_point, ] - 
                       getBiomass(model_results$species_5_only)[1, ]) / 
                       getBiomass(model_results$species_5_only)[1, ]
    
    wbream_decrease <- (getBiomass(model_results$species_4_only)[time_point, ] - 
                        getBiomass(model_results$species_4_only)[1, ]) / 
                        getBiomass(model_results$species_4_only)[1, ]
    
    bream_decrease <- (getBiomass(model_results$species_3_only)[time_point, ] - 
                       getBiomass(model_results$species_3_only)[1, ]) / 
                       getBiomass(model_results$species_3_only)[1, ]
    
    ruffe_decrease <- (getBiomass(model_results$species_2_only)[time_point, ] - 
                       getBiomass(model_results$species_2_only)[1, ]) / 
                       getBiomass(model_results$species_2_only)[1, ]
    
    small_fish_decrease <- (getBiomass(model_results$species_1_only)[time_point, ] - 
                            getBiomass(model_results$species_1_only)[1, ]) / 
                            getBiomass(model_results$species_1_only)[1, ]
    
    no_change <- (getBiomass(model_results$no_change)[time_point, ] - 
                  getBiomass(model_results$no_change)[1, ]) / 
                  getBiomass(model_results$no_change)[1, ]
    
    # Combine into matrix
    model_change_matrix <- rbind(
      small_fish_decrease, 
      ruffe_decrease, 
      bream_decrease, 
      wbream_decrease, 
      roach_decrease, 
      perch_decrease, 
      pikeperch_decrease, 
      five_species_decrease, 
      all_decrease, 
      no_change
    )
    
    # Store with model name
    model_changes[[model_name]] <- model_change_matrix
  }
  
  return(model_changes)
}

# Calculate changes for all models
all_model_changes <- calculate_change_all_models(all_model_results, time_point = 50)

# Convert to long format with all models
changes_long_all <- data.frame()

for (model_name in names(all_model_changes)) {
  change_matrix <- all_model_changes[[model_name]] * 100  # Convert to percent
  change_long <- melt(change_matrix)
  colnames(change_long) <- c("Scenario", "Species", "Change_percent")
  change_long$Model <- model_name
  changes_long_all <- rbind(changes_long_all, change_long)
}

# Calculate summary statistics with MIN/MAX
changes_summary <- changes_long_all %>%
  group_by(Scenario, Species) %>%
  summarise(
    mean_change = mean(Change_percent),
    median_change = median(Change_percent),
    min_change = min(Change_percent),      # Minimum across models
    max_change = max(Change_percent),      # Maximum across models
    range_width = max(Change_percent) - min(Change_percent),  # Width of range
    .groups = 'drop'
  )

# Better scenario labels
changes_summary <- changes_summary %>%
  mutate(Scenario_label = case_when(
    Scenario == "small_fish_decrease" ~ "small fish",
    Scenario == "ruffe_decrease" ~ "ruffe",
    Scenario == "bream_decrease" ~ "bream",
    Scenario == "wbream_decrease" ~ "white bream",
    Scenario == "roach_decrease" ~ "roach",
    Scenario == "perch_decrease" ~ "perch",
    Scenario == "pikeperch_decrease" ~ "pikeperch",
    Scenario == "five_species_decrease" ~ "5 commercial spp.",
    Scenario == "all_decrease" ~ "all 7 spp.",
    Scenario == "no_change" ~ "no change",
    TRUE ~ as.character(Scenario)
  ))

# Reorder scenarios
changes_summary$Scenario_label <- factor(changes_summary$Scenario_label,
                                    levels = c("No change", "small fish", "ruffe", 
                                              "bream", "white bream", "roach", 
                                               "perch", "pikeperch", "5 commercial spp.", 
                                                     "all 7 spp."))
```

```{r}
# After running no-fishing scenarios, save the summary
changes_summary_no_fishing <- changes_summary %>%
  filter(Scenario_label != "No change") %>%
  mutate(
    Fishing = "Without fishing",
    label_text = case_when(
      round(min_change, 0) == round(max_change, 0) ~ sprintf("%.0f", min_change),
      TRUE ~ sprintf("%.0f – %.0f", min_change, max_change)
    )
  )
```

```{r}
# Filter out "No change" scenario and create custom label
changes_summary <- changes_summary_no_fishing %>%
  filter(Scenario_label != "No change") %>%
  mutate(
    min_r = round(min_change, 0),
    max_r = round(max_change, 0),
    min_r = ifelse(min_r == 0, 0, min_r),
    max_r = ifelse(max_r == 0, 0, max_r),
    label_text = case_when(
      min_r == max_r ~ sprintf("%.0f", min_r),
      TRUE ~ sprintf("%.0f – %.0f", min_r, max_r)
    )
  ) %>%
  select(-min_r, -max_r)

# Reorder scenarios (without "No change")
changes_summary$Scenario_label <- factor(changes_summary$Scenario_label,
                                  levels = c("small fish", "ruffe", 
                                             "bream", "white bream", "roach", 
                                             "perch", "pikeperch", "5 commercial spp.", 
                                                     "all 7 spp."))

# Create plot
p_no_fising <- ggplot(changes_summary, aes(x = Species, y = Scenario_label, fill = mean_change)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = label_text), size = 3.5, lineheight = 1.1) +  # Increased size slightly
  scale_fill_gradient2(low = "#56b4e9", mid = "white", high = "#E69F00", 
                       midpoint = 0, name = "Mean\nChange (%)") +
  geom_hline(yintercept = 7.5, linewidth = 1.5, color = "grey30") +
  labs(x = "Species", 
       y = "w_mat Decrease Scenario",
       title = "Biomass Change at Year 50 (%)",
       subtitle = "Range (min – max) across 5 model parameterizations") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 11, face = "bold"),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 9, hjust = 0.5),
    legend.title = element_text(size = 9, face = "bold"),
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA)
  )

```

Combine both datasets with fishing and without

```{r}
# Combine both datasets
changes_summary_combined <- bind_rows(
  changes_summary_fishing,
  changes_summary_no_fishing
)

# Make sure Scenario_label levels are consistent
changes_summary_combined$Scenario_label <- factor(
  changes_summary_combined$Scenario_label,
  levels = c("small fish", "ruffe", "bream", "white bream", 
             "roach", "perch", "pikeperch", "5 commercial spp.", "all 7 spp.")
)

# Set fishing as a factor with desired order
changes_summary_combined$Fishing <- factor(
  changes_summary_combined$Fishing,
  levels = c("With fishing", "Without fishing")
)
```

```{r}
p_combined <- ggplot(changes_summary_combined, 
                     aes(x = Species, y = Scenario_label, fill = mean_change)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = label_text), size = 3) +
  scale_fill_gradient2(low = "#56b4e9", mid = "white", high = "#E69F00", 
                       midpoint = 0, name = "Mean change (%)") +
  scale_x_discrete(labels = c("small_fish" = "small fish", "white_bream" = "white bream")) +
  annotate("rect", 
           xmin = 0.5, xmax = length(unique(changes_summary_combined$Species)) + 0.5,
           ymin = 7.5, ymax = 8.5,  # adjust ymax to number of total levels + 0.5
           fill = NA, color = "grey70", linewidth = 1.2) +
  facet_wrap(~ Fishing, ncol = 2) +
  labs(x = "Species", 
       y = "Maturity size decrease scenario") +
  theme(
    axis.text.x = element_text(angle = 50, hjust = 1, size = 13),
    axis.text.y = element_text(size = 13),
    axis.title = element_text(size = 13, face = "bold"),
    plot.title = element_text(size = 13, face = "bold", hjust = 0.5),
    strip.text = element_text(size = 13, face = "bold"),
    strip.background = element_rect(fill = "grey90", color = NA),
    legend.title = element_text(size = 13, face = "bold"),
    legend.position = "bottom",
    panel.grid = element_blank(),
    plot.background = element_rect(fill = "white", color = NA)
  )

# ggsave("biomass_change_comparison.tiff", p_combined, width = 9, height = 7, dpi = 600, bg = "white")
```

size spectrum without fishing

```{r}
plotSpectra(all_model_results$M7_kappa9_lambda2.05extMort01_0.333.rds$no_change)
plotSpectra(all_model_results$M12_kappa7_lambda2.05extMort02_0.333.rds$no_change)
plotSpectra(all_model_results$M5_kappa7_lambda2.05extMort01_0.333.rds$no_change)
plotSpectra(all_model_results$M4_kappa7_lambda2.05extMort01_0.25.rds$no_change)
plotSpectra(all_model_results$M6_kappa8_lambda2.05extMort01_0.333.rds$no_change)
```

# diet change

For each predator body size class and time step, we calculated the percent change in proportional diet composition relative to the initial model state (year 1), allowing us to track how prey contributions to pikeperch diet shifted throughout the simulation in response to decreasing maturity size.

with ci

```{r}
# ── Extract diet data ──
diet_all_timesteps_list <- list()
for (model_name in names(all_model_results)) {
  for (scenario_name in names(all_model_results[[model_name]])) {
    
    sim    <- all_model_results[[model_name]][[scenario_name]]
    params <- sim@params
    times  <- getTimes(sim)
    
    diet_baseline_df <- reshape2::melt(
      getDiet(params, n = sim@n[1,,], n_pp = sim@n_pp[1,], n_other = sim@n_other[1,], proportion = TRUE),
      value.name = "baseline_value"
    )
    colnames(diet_baseline_df) <- c("Predator", "w", "Prey", "baseline_value")
    
    last_t <- length(times)
    diet_last_df <- reshape2::melt(
      getDiet(params, n = sim@n[last_t,,], n_pp = sim@n_pp[last_t,], n_other = sim@n_other[last_t,], proportion = TRUE),
      value.name = "current_value"
    )
    colnames(diet_last_df) <- c("Predator", "w", "Prey", "current_value")
    
    diet_change <- diet_baseline_df %>%
      left_join(diet_last_df, by = c("Predator", "w", "Prey")) %>%
      mutate(
        pp_change = (current_value - baseline_value) * 100,
        Scenario  = scenario_name,
        Model     = model_name
      )
    
    diet_all_timesteps_list[[paste0(model_name, "_", scenario_name)]] <- diet_change
  }
}
diet_all_timesteps_combined <- bind_rows(diet_all_timesteps_list)

# ── Filter for plotting — all 5 models, target scenario ──
target_scenario <- "species_3_4_5_6_7"

diet_plot_data <- diet_all_timesteps_combined %>%
  filter(
    Scenario == target_scenario,
    Predator %in% c("pikeperch", "perch"),
    Prey != "External"
  )

# ── Per-model summary: total pp_change per model × predator × prey ──
prey_summary_per_model <- diet_plot_data %>%
  group_by(Model, Predator, Prey) %>%
  summarise(
    total_baseline = sum(baseline_value, na.rm = TRUE),
    total_current  = sum(current_value,  na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(total_pp_change = (total_current - total_baseline) * 100)

# ── Interval summary: min–max across all 5 models per predator × prey ──
prey_summary_interval <- prey_summary_per_model %>%
  group_by(Predator, Prey) %>%
  summarise(
    min_change = min(total_pp_change, na.rm = TRUE),
    max_change = max(total_pp_change, na.rm = TRUE),
    .groups = "drop"
  ) %>%
 mutate(
    label = case_when(
      round(min_change, 0) == round(max_change, 0) ~
        paste0(ifelse(min_change > 0, "+", ""), round(min_change, 0), "%"),
      TRUE ~
        paste0(ifelse(min_change > 0, "+", ""), round(min_change, 0), " – ",
               ifelse(max_change > 0, "+", ""), round(max_change, 0), "%")
    ),
    mid_change  = (min_change + max_change) / 2,
    label_color = ifelse(mid_change > 0, "#b2182b", "black"),
    w = Inf
  )

# ── Mean across all 5 models for tile colours ──
diet_plot_data_fill <- diet_plot_data %>%
  group_by(w, Predator, Prey) %>%
  summarise(pp_change = mean(pp_change, na.rm = TRUE), .groups = "drop")

# ── Plot ──
diet <- ggplot(diet_plot_data_fill, aes(x = w, y = Prey, fill = pp_change)) +
  geom_tile(width = diff(range(log10(diet_plot_data_fill$w))) / 50) +
  geom_text(
    data        = prey_summary_interval,
    aes(x = w, y = Prey, label = label),
    inherit.aes = FALSE,
    hjust       = -0.05,
    size        = 3.5,
    fontface    = "bold",
    color       = prey_summary_interval$label_color
  ) +
  facet_wrap(~ Predator, scales = "free_y", ncol = 1) +
  scale_y_discrete(
  limits = c("pikeperch", "perch", "roach", "white_bream", "bream", "ruffe",  "small_fish", "Resource"),
  labels = function(x) gsub("_", " ", ifelse(x == "Resource", "resource", x))
) + 
  scale_x_log10(limits = c(5, NA), breaks = c(1, 10, 100, 1000)) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "#f7f7f7",
    high     = "#b2182b",
    midpoint = 0,
    limits   = c(-3, 3),
    na.value = "#f7f7f7",
    oob      = scales::squish,
    name     = "Change in diet proportion (%)"
  ) +
  labs(
    x        = "Predator weight (g)",
    y        = "Prey species"
  ) +
  My_theme +
  theme(
    legend.position = "bottom",
    axis.text.y     = element_text(angle = 0),
    strip.text      = element_text(face = "bold"),
    plot.margin     = margin(5, 90, 5, 5)
  ) +
  coord_cartesian(clip = "off")

# First, I run the simulation with fishing and save it as dietF, then rename it to diet when running without fishing.
dietF | diet

diet_clean <- diet + 
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())

combined_diet <- (dietF | diet_clean) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a") &
  theme(legend.position  = "bottom",
        legend.box       = "horizontal")

diet_titled       <- dietF       + ggtitle("With fishing")
diet_clean_titled <- diet_clean + ggtitle("Without fishing")

combined_diet <- (diet_titled | diet_clean_titled) +
  plot_layout(guides = "collect") &
  theme(
    legend.position  = "bottom",
    legend.box       = "horizontal",
    plot.title       = element_text(hjust = 0.5, face = "bold", size = 14) 
  )

combined_diet
# ggsave("combined_diet_change_mean.tiff", combined_diet, width = 9, height = 7, dpi = 600, bg = "white")
# ggsave("diet_change.tiff", diet, width = 6, height = 7, dpi = 600, bg = "white")
```

# Death Rate changes

```{r}
death_all_timesteps_list <- list()

for (model_name in names(all_model_results)) {
  for (scenario_name in names(all_model_results[[model_name]])) {
    
    sim         <- all_model_results[[model_name]][[scenario_name]]
    params      <- sim@params
    times       <- getTimes(sim)
    all_species <- params@species_params$species
    
    get_mort_df <- function(t) {
      pred_rate <- getPredRate(params, n = sim@n[t,,], n_pp = sim@n_pp[t,], n_other = sim@n_other[t,])
      f_mort    <- getFMort(params, effort = sim@effort[t,])
      
      do.call(rbind, lapply(all_species, function(sp) {
        idx_full <- params@w_full >= params@species_params[sp, "w_min"] &
                    params@w_full <= params@species_params[sp, "w_max"]
        idx      <- params@w      >= params@species_params[sp, "w_min"] &
                    params@w      <= params@species_params[sp, "w_max"]
        w_vals   <- params@w[idx]
        rbind(
          data.frame(w = w_vals, value = params@mu_b[sp, idx],           Cause = "External", Prey = sp),
          data.frame(w = w_vals, value = f_mort[sp, idx],                Cause = "Fishing",  Prey = sp),
          data.frame(w     = rep(w_vals, each = nrow(pred_rate)),
                     value = c(params@interaction[, sp] * pred_rate[, idx_full]),
                     Cause = rep(all_species, times = sum(idx)),
                     Prey  = sp)
        )
      }))
    }
    
    # ── Baseline (t = 1) ──
    baseline_df <- get_mort_df(1) %>% rename(baseline_value = value)
    
    # Total baseline predation summed across sizes per predator × prey
    # Computed from t=1 only — one row per predator × prey, no time inflation
    size_total_baseline <- baseline_df %>%
      filter(!Cause %in% c("Fishing", "External")) %>%
      group_by(Cause, Prey) %>%
      summarise(total_baseline = sum(baseline_value, na.rm = TRUE), .groups = "drop")
    
    # ── All time steps ──
    all_times_df <- do.call(rbind, lapply(seq_along(times), function(t) {
      get_mort_df(t) %>% mutate(time_step = times[t], time_index = t)
    })) %>% rename(current_value = value)
    
    # ── Join and normalize ──
    death_all_timesteps_list[[paste0(model_name, "_", scenario_name)]] <- all_times_df %>%
      left_join(baseline_df %>% select(w, Cause, Prey, baseline_value), by = c("w", "Cause", "Prey")) %>%
      left_join(size_total_baseline, by = c("Cause", "Prey")) %>%
      mutate(
        absolute_change   = current_value - baseline_value,
        normalized_change = ifelse(total_baseline > 0.001,
                                   (absolute_change / total_baseline) * 100,
                                   NA),
        Scenario = scenario_name,
        Model    = model_name
      )
  }
}

death_all_timesteps_combined <- bind_rows(death_all_timesteps_list)

```

with CI
```{r}
# ── Filter for all 5 models ──
target_scenario <- "species_3_4_5_6_7"
max_t           <- max(death_all_timesteps_combined$time_index)

death_heatmap_all_models <- death_all_timesteps_combined %>%
  filter(
    Scenario   == target_scenario,
    !Cause     %in% c("Fishing", "External", "small_fish", "ruffe"),
    time_index == max_t,
    baseline_value > 0.001
  )


# ── Per-model summary: total % change per model × prey ──
prey_label_per_model <- death_heatmap_all_models %>%
  group_by(Model, Prey) %>%
  summarise(
    total_absolute_change = sum(absolute_change,  na.rm = TRUE),
    total_baseline        = sum(baseline_value,   na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(pct_change = (total_absolute_change / total_baseline) * 100)

# ── Interval summary: min–max across all 5 models per prey ──
prey_interval_label <- prey_label_per_model %>%
  group_by(Prey) %>%
  summarise(
    min_change = min(pct_change, na.rm = TRUE),
    max_change = max(pct_change, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    label = case_when(
      round(min_change, 0) == round(max_change, 0) ~
        paste0(ifelse(min_change > 0, "+", ""), round(min_change, 0), "%"),
      TRUE ~
        paste0(ifelse(min_change > 0, "+", ""), round(min_change, 0), " ─ ",
               ifelse(max_change > 0, "+", ""), round(max_change, 0), "%")
    ),
    mid_change  = (min_change + max_change) / 2,
    label_color = ifelse(mid_change > 0, "#b2182b", "black"),
    y_pos = Inf
  )


# For each model, Prey, Cause — find the w where normalized_change crosses zero
zero_crossings <- death_heatmap_all_models %>%
  group_by(Model, Prey, Cause) %>%
  arrange(w) %>%
  mutate(sign_change = sign(normalized_change) != sign(lag(normalized_change))) %>%
  filter(sign_change) %>%
  summarise(zero_w = first(w), .groups = "drop")

# Summarise across models
zero_band <- zero_crossings %>%
  group_by(Prey, Cause) %>%
  summarise(
    zero_mean = mean(zero_w,   na.rm = TRUE),
    zero_min  = min(zero_w,    na.rm = TRUE),
    zero_max  = max(zero_w,    na.rm = TRUE),
    .groups = "drop"
  )

# Mean across all 5 models for tile colours
death_heatmap_final <- death_heatmap_all_models %>%
  group_by(w, Cause, Prey) %>%
  summarise(normalized_change = mean(normalized_change, na.rm = TRUE), .groups = "drop")

prey_order <- c("small_fish", "ruffe", "bream", "white_bream",
                "roach", "perch", "pikeperch")   # your full order, raw names

death_heatmap_final$Prey  <- factor(death_heatmap_final$Prey,  levels = prey_order)
prey_interval_label$Prey  <- factor(prey_interval_label$Prey,  levels = prey_order)
zero_band$Prey            <- factor(zero_band$Prey,            levels = prey_order)


death_rate <- ggplot(death_heatmap_final, aes(x = w, y = Cause, fill = normalized_change)) +
  geom_tile() +
  # Uncertainty band around zero-crossing
  geom_rect(
    data = zero_band,
    aes(xmin = zero_min, xmax = zero_max,
        ymin = as.numeric(factor(Cause)) - 0.5,
        ymax = as.numeric(factor(Cause)) + 0.5),
    inherit.aes = FALSE,
    fill  = NA,
    color = "black",
    linewidth = 0.5,
    linetype = "dashed"
  ) +
  # Mean zero-crossing line
  geom_segment(
    data = zero_band,
    aes(x = zero_mean, xend = zero_mean,
        y = as.numeric(factor(Cause)) - 0.5,
        yend = as.numeric(factor(Cause)) + 0.5),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 0.7
  ) +
  geom_text(
    data        = prey_interval_label,
    aes(x = 1000, y = y_pos, label = label),
    inherit.aes = FALSE,
    vjust       = 1.5,
    hjust       = 0,
    size        = 3.7,
    fontface    = "bold",
    color       = prey_interval_label$label_color
  ) +
  facet_wrap(~ Prey, scales = "fixed", ncol = 1,
           labeller = as_labeller(function(x) gsub("_", " ", x))) +
  scale_x_log10(
    limits = c(NA, 1000),
    breaks = c(0.1, 1, 10, 100, 1000),
    expand = expansion(mult = c(0.02, 0.15))
  ) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "#f7f7f7",
    high     = "#b2182b",
    midpoint = 0,
    limits   = c(-0.3, 0.3),
    na.value = "grey90",
    oob      = scales::squish,
    name     = "Mortality rate \nchange(%)"
  ) +
  labs(
    x     = "Prey weight (g)",
    y     = "Predator species",
    title = "Prey species"
  ) +
  My_theme +
  theme(
    plot.title      = element_text(hjust = 0.5, size = 15, family = "gillsans"),
    legend.position = "right",
    strip.text      = element_text(size = 10)
    ) +
  coord_cartesian(clip = "off")


# ggsave("mortality rate CI change.tiff", death_rate, width = 8, height = 6, dpi = 600, bg = "white")
```

### Yield changes

calibration with time series yields
(only works after scenario with fishing)
```{r}
# === 1. LOAD MODELS ===
models <- list(
  M3  = read_rds("M6_kappa8_lambda2.05extMort01_0.333.rds"),
  M4  = read_rds("M4_kappa7_lambda2.05extMort01_0.25.rds"),
  M5  = read_rds("M5_kappa7_lambda2.05extMort01_0.333.rds"),
  M12 = read_rds("M12_kappa7_lambda2.05extMort02_0.333.rds"),
  M13 = read_rds("M7_kappa9_lambda2.05extMort01_0.333.rds")
)


f_data        <- read.csv("../Data/CLF_1990_2023matrix_ci.csv")
years_data    <- f_data$Year
main_species  <- c("Bream", "Perch", "Pikeperch", "Roach", "White_bream")
f_main        <- as.matrix(f_data[, main_species])
rownames(f_main) <- as.character(years_data)
colnames(f_main) <- tolower(main_species)


add_ramp_up <- function(effort_matrix) {
  firstRecordedEffort <- effort_matrix[1, ]
  ramp_years  <- 1893:1991
  data_years  <- as.numeric(rownames(effort_matrix))
  initial_effort1 <- matrix(0, nrow = length(ramp_years), ncol = ncol(effort_matrix),
                             dimnames = list(as.character(ramp_years), colnames(effort_matrix)))
  for (i in 1:ncol(effort_matrix)) {
    initial_effort1[, i] <- seq(from = 0, to = firstRecordedEffort[i],
                                length.out = nrow(initial_effort1))
  }
  combined_effort <- rbind(initial_effort1, effort_matrix)
  rownames(combined_effort) <- as.character(c(ramp_years, data_years))
  return(combined_effort)
}


### delete -  old version with constant effort
run_model_simulation <- function(params, model_name) {
  current_gear        <- gear_params(params)
  current_gear$gear   <- current_gear$species
  gear_params(params) <- current_gear
  params <- setParams(params)

  f_main_complete <- cbind(small_fish = rep(0, nrow(f_main)),
                           ruffe      = rep(0, nrow(f_main)),
                           f_main)
  rownames(f_main_complete) <- as.character(years_data)

  sim2 <- project(params, effort = initial_effort, t_max = 300, dt = 0.1)

  species_order     <- species_params(sim2@params)$species
  f_main_ordered    <- f_main_complete[, species_order]
  relative_effort   <- sweep(f_main_ordered, 2, gear_params(sim2@params)$catchability, "/")
  relative_effort[is.nan(relative_effort)] <- 0

  relative_effort_full <- add_ramp_up(relative_effort)

  sim3 <- project(sim2@params, effort = relative_effort_full,
                  dt = 0.1, t_save = 1,
                  initial_n   = sim2@n[300, , ],
                  initial_npp = sim2@n_pp[300, ])

  y <- getYield(sim3) %>%
    melt() %>%
    mutate(model = model_name)
  return(y)
}

y_all_models <- lapply(names(models), function(name) {
  run_model_simulation(models[[name]], name)
}) %>% bind_rows()

y_all_models_filtered <- y_all_models %>%
  filter(!sp %in% c("small_fish", "ruffe"))

# === 2. GET BASELINE & MATURITY DECLINE YIELDS ===
# NOTE: all_model_results must already be loaded in your environment
y_baseline <- data.frame()
for (model_name in names(all_model_results)) {
  sim       <- all_model_results[[model_name]]$no_change
  yield_data <- getYield(sim) %>%
    melt() %>%
    mutate(model    = gsub(".rds", "", model_name),
           scenario = "Baseline")
  y_baseline <- bind_rows(y_baseline, yield_data)
}

y_maturity_decline <- data.frame()
for (model_name in names(all_model_results)) {
  sim       <- all_model_results[[model_name]]$species_3_4_5_6_7
  yield_data <- getYield(sim) %>%
    melt() %>%
    mutate(model    = gsub(".rds", "", model_name),
           scenario = "With maturity decline")
  y_maturity_decline <- bind_rows(y_maturity_decline, yield_data)
}

y_baseline         <- y_baseline         %>% filter(!sp %in% c("small_fish", "ruffe"))
y_maturity_decline <- y_maturity_decline %>% filter(!sp %in% c("small_fish", "ruffe"))

# === 3. MULTIPLIER EXTRACTION & APPLICATION ===
y_baseline_1_30 <- y_baseline %>% filter(time >= 1 & time <= 30)

y_curved_remapped <- y_all_models_filtered %>%
  filter(time >= 1993 & time <= 2022) %>%
  mutate(time_remapped = time - 1992)

curved_values <- y_curved_remapped %>%
  select(-time) %>%
  rename(time = time_remapped) %>%
  arrange(model, sp, time)

straight_values <- y_baseline_1_30 %>% arrange(model, sp, time)

extract_all_multipliers <- function(straight_data, curved_data) {
  straight_models <- unique(straight_data$model)
  curved_models   <- unique(curved_data$model)
  species_list    <- unique(straight_data$sp)
  all_multipliers <- data.frame()

  for (sp in species_list) {
    for (i in seq_along(straight_models)) {
      straight_model <- straight_models[i]
      if (i <= length(curved_models)) {
        curved_model <- curved_models[i]

        straight_vals <- straight_data %>%
          filter(model == straight_model, sp == !!sp) %>%
          pull(value)

        curved_vals <- curved_data %>%
          filter(model == curved_model, sp == !!sp) %>%
          pull(value)

        if (length(straight_vals) > 0 && length(curved_vals) > 0) {
          straight_constant <- straight_vals[1]
          multipliers       <- curved_vals / straight_constant

          multiplier_data <- data.frame(
            model_straight = straight_model,
            model_curved   = curved_model,
            species        = sp,
            time           = 1:length(multipliers),
            straight_value = straight_constant,
            curved_value   = curved_vals,
            multiplier     = multipliers
          )
          all_multipliers <- rbind(all_multipliers, multiplier_data)
        }
      }
    }
  }
  return(all_multipliers)
}

apply_all_multipliers <- function(straight_data, multiplier_table) {
  straight_data %>%
    left_join(multiplier_table %>% select(model_straight, species, time, multiplier),
              by = c("model" = "model_straight", "sp" = "species", "time" = "time")) %>%
    mutate(value_curved = value * ifelse(is.na(multiplier), 1.0, multiplier)) %>%
    select(-multiplier, -value) %>%
    rename(value = value_curved)
}

all_multipliers <- extract_all_multipliers(straight_values, curved_values)

y_baseline_curved  <- apply_all_multipliers(y_baseline %>% filter(time <= 30), all_multipliers) %>%
  mutate(scenario = "Baseline")
y_maturity_curved  <- apply_all_multipliers(y_maturity_decline %>% filter(time <= 30), all_multipliers) %>%
  mutate(scenario = "With maturity decline")
y_final_curved     <- bind_rows(y_baseline_curved, y_maturity_curved)

# === 4. REVERSED YIELD CHANGES ===
yield_changes_reversed <- y_final_curved %>%
  select(time, sp, model, scenario, value) %>%
  pivot_wider(names_from = scenario, values_from = value) %>%
  mutate(
    absolute_change = `With maturity decline` - Baseline,
    percent_change  = ((`With maturity decline` - Baseline) / Baseline) * 100,
    year            = 1992 + time
  )

uncertainty_analysis_reversed <- yield_changes_reversed %>%
  group_by(sp, time, year) %>%
  summarise(
    mean_percent_change = mean(percent_change, na.rm = TRUE),
    min_percent_change  = min(percent_change,  na.rm = TRUE),
    max_percent_change  = max(percent_change,  na.rm = TRUE),
    n_models            = n(),
    .groups = "drop"
  )

# === 5. CUMULATIVE LOSS LABELS ===
cumulative_loss_labels <- uncertainty_analysis_reversed %>%
  group_by(sp) %>%
  summarise(
    total_min_loss = sum(min_percent_change, na.rm = TRUE),
    total_max_loss = sum(max_percent_change, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("Cumulative loss: ", round(total_max_loss, 0), " — ", round(total_min_loss, 0), "%"),
    x_pos = 2000,
    y_pos = -15
  )

# === 6. FINAL PLOT ===
yield_loss <- ggplot() +
  geom_ribbon(data = uncertainty_analysis_reversed,
              aes(x = year, ymin = min_percent_change, ymax = max_percent_change),
              fill = "grey70", alpha = 0.4) +
  geom_line(data = yield_changes_reversed,
            aes(x = year, y = percent_change, group = model),
            color = "grey20", alpha = 0.6, linewidth = 0.7) +
  geom_hline(yintercept = 0,    linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = -Inf, color = "gray50", linewidth = 0.3) +
  geom_hline(yintercept = -Inf, color = "gray50", linewidth = 0.3) +
  geom_text(data = cumulative_loss_labels,
            aes(x = x_pos, y = y_pos, label = label),
            size = 4, fontface = "bold", hjust = 0, nudge_x = -8,
            inherit.aes = FALSE) +
  facet_wrap(~sp, scales = "fixed", ncol = 2) +
  scale_y_continuous(limits = c(-17, 0), breaks = seq(-15, 0, by = 5)) +
  labs(x = "Time (year)", y = "Percent Change (%)") +
  My_theme

# ggsave("yield_loss.tiff", yield_loss, width = 7, height = 7, dpi = 600, bg = "white")
```

```{r}
# === PERCENT-CHANGE PLOT (single-colour version of the orange/blue figure) ===
# Uses y_final_curved from the multiplier workflow (curved data, time 1–30)

# --- 1. Percent change per model, per species, per timestep ---
yield_changes <- y_final_curved %>%
  select(time, sp, model, scenario, value) %>%
  pivot_wider(names_from = scenario, values_from = value) %>%
  mutate(
    percent_change = ((`With maturity decline` - Baseline) / Baseline) * 100,
    year           = 1992 + time          # 1 -> 1993, 30 -> 2022
  )

# --- 2. Across-model min–max range (the shaded background) ---
yield_range <- yield_changes %>%
  group_by(sp, time, year) %>%
  summarise(
    min_percent_change = min(percent_change, na.rm = TRUE),
    max_percent_change = max(percent_change, na.rm = TRUE),
    .groups = "drop"
  )

# --- 3. Cumulative loss label per species (summed across all years) ---
cumulative_loss_labels <- yield_range %>%
  group_by(sp) %>%
  summarise(
    total_min_loss = sum(min_percent_change, na.rm = TRUE),
    total_max_loss = sum(max_percent_change, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("Cumulative change: ",
                   round(total_max_loss, 0), " – ",
                   round(total_min_loss, 0), "%")
  )

# --- 4. Plot ---
yield_loss <- ggplot() +
  # shaded across-model range
  geom_ribbon(data = yield_range,
              aes(x = year, ymin = min_percent_change, ymax = max_percent_change),
              fill = "grey70", alpha = 0.4) +
  # individual model lines
  geom_line(data = yield_changes,
            aes(x = year, y = percent_change, group = model),
            colour = "grey20", alpha = 0.6, linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "gray50") +
  # cumulative loss label (top-left of each panel)
  geom_text(data = cumulative_loss_labels,
            aes(x = 1993, y = Inf, label = label),
            size = 3.5, fontface = "bold",
            hjust = 0, vjust = 1.5,
            inherit.aes = FALSE) +
  facet_wrap(~sp, scales = "free_y") +
  scale_x_continuous(breaks = c(1993, 2002, 2012, 2022),
                     labels = c("1993", "2002", "2012", "2022")) +
  labs(x = "Year",
       y = "Yield change (%)") +
  My_theme +
  theme(
    panel.border    = element_blank(),
    axis.line.x     = element_line(color = "gray50", linewidth = 0.3),
    axis.line.y     = element_line(color = "gray50", linewidth = 0.3),
    panel.spacing   = unit(1, "lines"),
    strip.background = element_blank(),
    strip.text      = element_text(face = "bold")
  )

yield_loss

```

# Size spectra change

```{r}
sf_list <- list()

for (model_name in names(all_model_results)) {
  model_results <- all_model_results[[model_name]]
  
  for (scenario_name in names(model_results)) {
    sim <- model_results[[scenario_name]]
    times <- getTimes(sim)
    first_time <- times[1]
    last_time  <- times[length(times)]
    
    # Get size class widths from the model
    dw <- sim@params@dw
    w  <- sim@params@w
    
    sf1 <- mizer::plotSpectra(sim, time_range = first_time, return_data = TRUE)
    sf2 <- mizer::plotSpectra(sim, time_range = last_time,  return_data = TRUE)
    
    # Join and convert density to actual numbers by multiplying by dw
    sf <- left_join(sf1, sf2, by = c("w", "Legend")) %>%
      left_join(data.frame(w = w, dw = dw), by = "w") %>%
      mutate(
        # Convert density to numbers
        N_first   = value.x * dw,
        N_last    = value.y * dw,
        # Percent change in actual numbers
        pct_change = ((N_last - N_first) / N_first) * 100,
        Scenario  = scenario_name,
        Model     = gsub(".rds", "", model_name)
      )
    
    sf_list[[paste(model_name, scenario_name, sep = "_")]] <- sf
  }
}

sf_combined <- bind_rows(sf_list)


```

```{r}
species_colors <- c(
  "small_fish" = "#FCC300",
  "ruffe" = "#44AA99", 
  "bream" = "#b2182b",
  "white_bream" = "#661100",
  "roach" = "#AA4499",
  "perch" = "#2166ac",
  "pikeperch" = "#93AA00"
)

# Option A: Percent change in actual numbers
ggplot(sf_combined %>% filter(Scenario == "species_3_4_5_6_7"),
       aes(x = w, y = pct_change,
           colour = Legend,
           group = interaction(Legend, Model))) +
  geom_line(linewidth = 1.3, alpha = 0.9) +
  geom_hline(yintercept = 0, linetype = 1, colour = "dark grey", linewidth = 0.75) +
  labs(x = "Weight [g]",
       y = "Percent change in abundance (%)",
       title = "Size spectrum change - Species 3-7 scenario",
       colour = "Species") +
  coord_cartesian(xlim = c(1, NA), ylim = c(-100, 150)) +
  scale_x_log10() +
  scale_colour_manual(values = species_colors) +
  My_theme +
  theme(legend.position = "bottom")

```

```{r}
# 1. Get w bins
w_bins <- all_model_results$`M5_kappa7_lambda2.05extMort01_0.333.rds`$all_species@params@w

# 2. Snap thresholds to nearest bin
w_max_final <- tibble(
  Legend       = c("small_fish", "ruffe", "bream", "white_bream", "roach", "perch", "pikeperch"),
  w_max_thresh = c(Inf, Inf, 2783.0279, 474.9298, 719.9572, 1107.1611, 3630.0364)
) %>%
  mutate(w_max_thresh = ifelse(is.infinite(w_max_thresh), Inf,
           w_bins[sapply(w_max_thresh, function(x) which.min(abs(w_bins - x)))]))

# 3. Species order
species_order <- c("pikeperch", "bream", "perch", "roach", "white_bream", "ruffe", "small_fish")

# 4. w_mat reference lines data
w_max_t50 <- data.frame(
  Legend        = species_params(cur_model)$species,
  w_max_initial = species_params(cur_model)$w_max,
  w_mat_initial = species_params(cur_model)$w_mat
) %>%
  mutate(
    decrease_rate = 0.01,
    w_max_t50 = w_max_initial * (1 - decrease_rate)^50,
    w_mat_t50 = w_mat_initial * (1 - decrease_rate)^50,
    w_max_t50 = ifelse(Legend %in% c("small_fish", "ruffe"), w_max_initial, w_max_t50),
    w_mat_t50 = ifelse(Legend %in% c("small_fish", "ruffe"), w_mat_initial, w_mat_t50)
  )

# 5. Build nd_ratio ONCE (used for both plots)
nd_ratio <- sf_combined %>%
  filter(Scenario == "species_3_4_5_6_7", N_first > 1e-10) %>%
  mutate(ratio = N_last / N_first) %>%
  left_join(w_max_final, by = "Legend") %>%
  mutate(reliable = w <= w_max_thresh)

# 6. Build nd_ratio_plot for the ratio plot
nd_ratio_plot <- bind_rows(
  nd_ratio,
  nd_ratio %>%
    group_by(Legend, Model) %>%
    slice(which(w == w_max_thresh)[1]) %>%
    mutate(reliable = FALSE)
) %>%
  arrange(Legend, Model, w) %>%
  mutate(Legend = factor(Legend, levels = species_order))

# 7. Build nd_ratio_p1_plot for the pct_change plot (same data, different y)
nd_ratio_p2_plot <- nd_ratio_plot  # identical structure, pct_change column already present

# 8. Plot ratio
ggplot() +
  geom_line(data = nd_ratio_plot %>% filter(reliable),
            aes(x = w, y = ratio, colour = Legend,
                group = interaction(Legend, Model)),
            linewidth = 0.8, alpha = 0.7, linetype = "solid") +
  geom_line(data = nd_ratio_plot %>% filter(!reliable),
            aes(x = w, y = ratio, colour = Legend,
                group = interaction(Legend, Model)),
            linewidth = 1, alpha = 0.6, linetype = "dotted") +
  geom_hline(yintercept = 1, colour = "grey50") +
  scale_x_log10() +
  scale_colour_manual(values = species_colors) +
  coord_cartesian(ylim = c(0, 3)) +
  labs(x = "Weight [g]", y = "N_last / N_first (ratio)") +
  My_theme +
  theme(legend.position = "bottom")

# 9. Plot pct_change (p1)
p2 <- ggplot() +
  geom_line(data = nd_ratio_p2_plot %>% filter(reliable),
            aes(x = w, y = pct_change, colour = Legend,
                group = interaction(Legend, Model)),
            linewidth = 0.8, alpha = 0.7, linetype = "solid") +
  geom_line(data = nd_ratio_p2_plot %>% filter(!reliable),
            aes(x = w, y = pct_change, colour = Legend,
                group = interaction(Legend, Model)),
            linewidth = 1, alpha = 0.6, linetype = "dotted") +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_vline(data = w_max_t50,
             aes(xintercept = w_mat_t50, colour = Legend),
             linetype = "dashed", linewidth = 0.7) +
  geom_vline(data = w_max_t50,
             aes(xintercept = w_mat_initial, colour = Legend),
             linetype = "solid", linewidth = 0.7) +
  scale_x_log10(breaks = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 4000),
                labels = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 4000)) +
  scale_colour_manual(values = species_colors) +
  coord_cartesian(ylim = c(-100, 200), xlim = c(1, NA)) +
  labs(x = "Weight (g)", y = "Change in abundance (%)", colour = "Species") +
  My_theme +
  theme(legend.position = "right")
```

```{r}
p2_inset <- ggplot(nd_ratio, aes(x = w, y = pct_change,
                     colour = Legend,
                     group = interaction(Legend, Model))) +
  geom_line(linewidth = 0.8, alpha = 0.7) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_vline(data = w_max_t50,
             aes(xintercept = w_mat_t50, colour = Legend),
             linetype = "dashed", linewidth = 0.7) +
  scale_x_log10(breaks = c(1e-03, 1e-01, 10),
              labels = c(1e-03, 1e-01, 10)) +
  scale_colour_manual(values = species_colors) +
  coord_cartesian(ylim = c(-20, 40), xlim = c(1e-03, 1e+01)) +
  labs(x = NULL, y = NULL) +
  My_theme +
  theme(
    legend.position = "none",
    plot.background = element_rect(fill = "white", colour = "black", linewidth = 0.2),
    panel.background = element_rect(fill = "white")
  )

# Combine with inset
spectrum2 <- p2 + inset_element(p2_inset, left = 0.01, bottom = 0.6, right = 0.35, top = 0.99)

# ggsave("size spectra1.tiff", spectrum, width = 9, height = 5, dpi = 600, bg = "white")
```

adding bar plot beneath

```{r}
# ── Step 1: rebuild p1 with correct x limits ──
p1 <- ggplot() +
  geom_line(data = nd_ratio_p1_plot %>% filter(reliable),
            aes(x = w, y = pct_change, colour = Legend,
                group = interaction(Legend, Model)),
            linewidth = 0.8, alpha = 0.7, linetype = "solid") +
  geom_line(data = nd_ratio_p1_plot %>% filter(!reliable),
            aes(x = w, y = pct_change, colour = Legend,
                group = interaction(Legend, Model)),
            linewidth = 1, alpha = 0.6, linetype = "dotted") +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_vline(data = w_max_t50,
             aes(xintercept = w_mat_t50, colour = Legend),
             linetype = "dashed", linewidth = 0.7) +
  geom_vline(data = w_max_t50,
             aes(xintercept = w_mat_initial, colour = Legend),
             linetype = "solid", linewidth = 0.7) +
  scale_x_log10(
    limits = c(1, 6000),
    breaks = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000),
    labels = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000)
  ) +
  scale_colour_manual(values = species_colors) +
  coord_cartesian(ylim = c(-100, 200)) +
  labs(x = NULL,                            # no x label — p2 carries it
       y = "Change in abundance (%)", 
       colour = "Species") +
  My_theme +
  theme(
    legend.position = "right",
    axis.text.x  = element_text(),          # keep tick labels on p1
    axis.ticks.x = element_line()
  )

# ── Step 2: attach inset to p1 ──
spectrum1 <- p1 + inset_element(p1_inset,    # your small inset plot (renamed to avoid clash)
                               left   = 0.01, 
                               bottom = 0.6, 
                               right  = 0.35, 
                               top    = 0.99)

# ── Bream-only heatmap (panel b) — ALL models ──
death_heatmap_bream_all <- death_all_timesteps_combined %>%
  filter(
    Scenario   == target_scenario,
    !Cause     %in% c("Fishing", "External", "small_fish", "ruffe"),
    time_index == max_t,
    baseline_value > 0.001,
    Prey       == "bream",
    w          >= 1
  ) 

# ── Mean across models for colouring ──
death_heatmap_bream <- death_heatmap_bream_all %>%
  group_by(w, Cause) %>%
  summarise(normalized_change = mean(normalized_change, na.rm = TRUE), .groups = "drop")

# ── Zero-crossing band across models ──
zero_band_bream <- death_heatmap_bream_all %>%
  group_by(Model, Cause) %>%
  arrange(w) %>%
  mutate(sign_change = sign(normalized_change) != sign(lag(normalized_change))) %>%
  filter(sign_change) %>%
  summarise(zero_w = first(w), .groups = "drop") %>%
  group_by(Cause) %>%
  summarise(
    zero_mean = mean(zero_w, na.rm = TRUE),
    zero_min  = min(zero_w,  na.rm = TRUE),
    zero_max  = max(zero_w,  na.rm = TRUE),
    .groups   = "drop"
  )

# ── Plot ──
p_heat1 <- ggplot(death_heatmap_bream, aes(x = w, y = Cause, fill = normalized_change)) +
  geom_tile() +
  
  # Uncertainty band (range of zero-crossings across models)
  geom_rect(
    data = zero_band_bream,
    aes(xmin = zero_min, xmax = zero_max,
        ymin = as.numeric(factor(Cause)) - 0.5,
        ymax = as.numeric(factor(Cause)) + 0.5),
    inherit.aes = FALSE,
    fill      = NA,
    color     = "black",
    linewidth = 0.5,
    linetype  = "dashed"
  ) +
  
  # Mean zero-crossing line
  geom_segment(
    data = zero_band_bream,
    aes(x    = zero_mean, xend = zero_mean,
        y    = as.numeric(factor(Cause)) - 0.5,
        yend = as.numeric(factor(Cause)) + 0.5),
    inherit.aes = FALSE,
    color     = "black",
    linewidth = 0.7
  ) +
  
  scale_x_log10(
    limits = c(1, 6000),
    breaks = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000),
    labels = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_fill_gradient2(
    low      = "#2166ac", mid = "#f7f7f7", high = "#b2182b",
    midpoint = 0, limits = c(-0.3, 0.3),
    breaks   = c(-0.2, -0.1, 0, 0.1, 0.2),
    labels   = c("-0.2", "-0.1", "0", "0.1", "0.2"),
    na.value = "grey90", oob = scales::squish,
    name     = "Predation mortality rate change (%)",
    guide    = guide_colorbar(
      title.position = "top",
      barwidth       = unit(8, "lines"),
      barheight      = unit(0.6, "lines"),
      ticks          = TRUE,
      label.theme    = element_text(size = 8)
    )
  ) +
  labs(x = "Weight (g)", y = "Predator") +
  My_theme +
  theme(
    legend.position  = "bottom",
    legend.direction = "horizontal",
    plot.margin      = margin(0, 5, 5, 5)
  ) +
  coord_cartesian(clip = "off")


# ── Recombine ──
final_plot1 <- spectrum1 / p_heat1 +
  plot_layout(heights = c(4, 0.5), guides = "keep") 
  # + plot_annotation(tag_levels = "a")

final_plot1

final_plot1 | final_plot2
# ggsave("size_spectra_with_heatmap.tiff", final_plot, width = 9, height = 7, dpi = 600, bg = "white")
```

common plot

```{r}
# ── Titles go on the spectrum (p1), not the final_plot ──
p1_fishing    <- p1 + ggtitle("With fishing") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "none")

p1_no_fishing <- p2 + ggtitle("Without Fishing") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
        legend.position = "none")

# ── Rebuild spectrums with titled p1 ──
spectrum_fishing    <- p1_fishing    + inset_element(p1_inset, left = 0.01, bottom = 0.6,
                                                      right = 0.35, top = 0.99)
spectrum_no_fishing <- p1_no_fishing + inset_element(p2_inset, left = 0.01, bottom = 0.5,
                                                      right = 0.45, top = 0.99)

# ── Rebuild final plots ──
final_plot1 <- spectrum_fishing    / p_heat1 + plot_layout(heights = c(4, 0.5), guides = "keep")
final_plot2 <- spectrum_no_fishing / p_heat2 + plot_layout(heights = c(4, 0.5), guides = "keep")

# ── Strip y-axis from right panel ──
final_plot2_clean <- final_plot2 & 
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())

# ── Shared legend from p1 ──
shared_legend <- get_legend(
  p1 + theme(legend.position  = "bottom",
             legend.direction = "horizontal")
)

# ── Combine panels ──
combined <- (final_plot1 | final_plot2_clean) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# ── Add shared legend below ──
final_with_legend <- plot_grid(
  patchworkGrob(combined),
  shared_legend,
  ncol        = 1,
  rel_heights = c(1, 0.05)
)

final_with_legend
```

```{r}
# ── p1 ──
p1 <- ggplot() +
  geom_line(data = nd_ratio_p1_plot %>% filter(reliable),
            aes(x = w, y = pct_change, colour = Legend,
                group = interaction(Legend, Model)),
            linewidth = 0.8, alpha = 0.7, linetype = "solid") +
  geom_line(data = nd_ratio_p1_plot %>% filter(!reliable),
            aes(x = w, y = pct_change, colour = Legend,
                group = interaction(Legend, Model)),
            linewidth = 1, alpha = 0.6, linetype = "dotted") +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_vline(data = w_max_t50,
             aes(xintercept = w_mat_t50, colour = Legend),
             linetype = "dashed", linewidth = 0.7) +
  geom_vline(data = w_max_t50,
             aes(xintercept = w_mat_initial, colour = Legend),
             linetype = "solid", linewidth = 0.7) +
  scale_x_log10(
    limits = c(1, 6000),
    breaks = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000),
    labels = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000)
  ) +
  scale_colour_manual(values = species_colors) +
  coord_cartesian(ylim = c(-100, 200), xlim = c(1, 6000), clip = "off") +
  labs(x = NULL, y = "Change in abundance (%)", colour = "Species") +
  My_theme +
  theme(legend.position = "none")  # removed here, added via shared legend later

# ── Inset ──
spectrum1 <- p1 + inset_element(p1_inset,
                                left = 0.01, bottom = 0.6,
                                right = 0.35, top = 0.99)

# ── Heatmap data ──
death_heatmap_bream_all <- death_all_timesteps_combined %>%
  filter(
    Scenario       == target_scenario,
    !Cause         %in% c("Fishing", "External", "small_fish", "ruffe"),
    time_index     == max_t,
    baseline_value > 0.001,
    Prey           == "bream",
    w              >= 1
  )

death_heatmap_bream <- death_heatmap_bream_all %>%
  group_by(w, Cause) %>%
  summarise(normalized_change = mean(normalized_change, na.rm = TRUE), .groups = "drop")

zero_band_bream <- death_heatmap_bream_all %>%
  group_by(Model, Cause) %>%
  arrange(w) %>%
  mutate(sign_change = sign(normalized_change) != sign(lag(normalized_change))) %>%
  filter(sign_change) %>%
  summarise(zero_w = first(w), .groups = "drop") %>%
  group_by(Cause) %>%
  summarise(
    zero_mean = mean(zero_w, na.rm = TRUE),
    zero_min  = min(zero_w,  na.rm = TRUE),
    zero_max  = max(zero_w,  na.rm = TRUE),
    .groups   = "drop"
  )

# ── p_heat1 — matching x scale to p1 ──
p_heat1 <- ggplot(death_heatmap_bream, aes(x = w, y = Cause, fill = normalized_change)) +
  geom_tile() +
  geom_rect(
    data = zero_band_bream,
    aes(xmin = zero_min, xmax = zero_max,
        ymin = as.numeric(factor(Cause)) - 0.5,
        ymax = as.numeric(factor(Cause)) + 0.5),
    inherit.aes = FALSE,
    fill = NA, color = "black", linewidth = 0.5, linetype = "dashed"
  ) +
  geom_segment(
    data = zero_band_bream,
    aes(x = zero_mean, xend = zero_mean,
        y = as.numeric(factor(Cause)) - 0.5,
        yend = as.numeric(factor(Cause)) + 0.5),
    inherit.aes = FALSE,
    color = "black", linewidth = 0.7
  ) +
  scale_x_log10(
    limits = c(1, 6000),
    breaks = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000),
    labels = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000)
    # no expand — matches p1 default
  ) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "#f7f7f7", high = "#b2182b",
    midpoint = 0, limits = c(-0.3, 0.3),
    breaks = c(-0.2, -0.1, 0, 0.1, 0.2),
    labels = c("-0.2", "-0.1", "0", "0.1", "0.2"),
    na.value = "grey90", oob = scales::squish,
    name = "Predation mortality rate change (%)",
    guide = guide_colorbar(
      title.position = "top",
      barwidth       = unit(8, "lines"),
      barheight      = unit(0.6, "lines"),
      ticks          = TRUE,
      label.theme    = element_text(size = 8)
    )
  ) +
  labs(x = "Weight (g)", y = "Predator") +
  My_theme +
  theme(
    legend.position  = "bottom",
    legend.direction = "horizontal",
    plot.margin      = margin(0, 5, 5, 5)
  ) +
  coord_cartesian(xlim = c(1, 6000), clip = "off")  # matches p1

# ── Combine ──
final_plot1 <- spectrum1 / p_heat1 +
  plot_layout(heights = c(4, 0.5), guides = "keep")

final_plot1
```

```{r}
# ── p1 ──
p2 <- ggplot() +
  geom_line(data = nd_ratio_p2_plot %>% filter(reliable),
            aes(x = w, y = pct_change, colour = Legend,
                group = interaction(Legend, Model)),
            linewidth = 0.8, alpha = 0.7, linetype = "solid") +
  geom_line(data = nd_ratio_p2_plot %>% filter(!reliable),
            aes(x = w, y = pct_change, colour = Legend,
                group = interaction(Legend, Model)),
            linewidth = 1, alpha = 0.6, linetype = "dotted") +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_vline(data = w_max_t50,
             aes(xintercept = w_mat_t50, colour = Legend),
             linetype = "dashed", linewidth = 0.7) +
  geom_vline(data = w_max_t50,
             aes(xintercept = w_mat_initial, colour = Legend),
             linetype = "solid", linewidth = 0.7) +
  scale_x_log10(
    limits = c(1, 6000),
    breaks = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000),
    labels = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000)
  ) +
  scale_colour_manual(values = species_colors) +
  coord_cartesian(ylim = c(-100, 200), xlim = c(1, 6000), clip = "off") +
  labs(x = NULL, y = "Change in abundance (%)", colour = "Species") +
  My_theme +
  theme(legend.position = "none")  # removed here, added via shared legend later

# ── Inset ──
spectrum2 <- p2 + inset_element(p2_inset,
                                left = 0.01, bottom = 0.6,
                                right = 0.35, top = 0.99)

# ── Heatmap data ──
death_heatmap_bream_all <- death_all_timesteps_combined %>%
  filter(
    Scenario       == target_scenario,
    !Cause         %in% c("Fishing", "External", "small_fish", "ruffe"),
    time_index     == max_t,
    baseline_value > 0.001,
    Prey           == "bream",
    w              >= 1
  )

death_heatmap_bream <- death_heatmap_bream_all %>%
  group_by(w, Cause) %>%
  summarise(normalized_change = mean(normalized_change, na.rm = TRUE), .groups = "drop")

zero_band_bream <- death_heatmap_bream_all %>%
  group_by(Model, Cause) %>%
  arrange(w) %>%
  mutate(sign_change = sign(normalized_change) != sign(lag(normalized_change))) %>%
  filter(sign_change) %>%
  summarise(zero_w = first(w), .groups = "drop") %>%
  group_by(Cause) %>%
  summarise(
    zero_mean = mean(zero_w, na.rm = TRUE),
    zero_min  = min(zero_w,  na.rm = TRUE),
    zero_max  = max(zero_w,  na.rm = TRUE),
    .groups   = "drop"
  )

# ── p_heat1 — matching x scale to p1 ──
p_heat2 <- ggplot(death_heatmap_bream, aes(x = w, y = Cause, fill = normalized_change)) +
  geom_tile() +
  geom_rect(
    data = zero_band_bream,
    aes(xmin = zero_min, xmax = zero_max,
        ymin = as.numeric(factor(Cause)) - 0.5,
        ymax = as.numeric(factor(Cause)) + 0.5),
    inherit.aes = FALSE,
    fill = NA, color = "black", linewidth = 0.5, linetype = "dashed"
  ) +
  geom_segment(
    data = zero_band_bream,
    aes(x = zero_mean, xend = zero_mean,
        y = as.numeric(factor(Cause)) - 0.5,
        yend = as.numeric(factor(Cause)) + 0.5),
    inherit.aes = FALSE,
    color = "black", linewidth = 0.7
  ) +
  scale_x_log10(
    limits = c(1, 6000),
    breaks = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000),
    labels = c(1, 3, 10, 30, 100, 300, 500, 1000, 2000, 6000)
    # no expand — matches p1 default
  ) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "#f7f7f7", high = "#b2182b",
    midpoint = 0, limits = c(-0.3, 0.3),
    breaks = c(-0.2, -0.1, 0, 0.1, 0.2),
    labels = c("-0.2", "-0.1", "0", "0.1", "0.2"),
    na.value = "grey90", oob = scales::squish,
    name = "Predation mortality rate change (%)",
    guide = guide_colorbar(
      title.position = "top",
      barwidth       = unit(8, "lines"),
      barheight      = unit(0.6, "lines"),
      ticks          = TRUE,
      label.theme    = element_text(size = 8)
    )
  ) +
  labs(x = "Weight (g)", y = "Predator") +
  My_theme +
  theme(
    legend.position  = "bottom",
    legend.direction = "horizontal",
    plot.margin      = margin(0, 5, 5, 5)
  ) +
  coord_cartesian(xlim = c(1, 6000), clip = "off")  # matches p1

# ── Combine ──
final_plot2 <- spectrum2 / p_heat2 +
  plot_layout(heights = c(4, 0.5), guides = "keep")

final_plot1 | final_plot2

library(cowplot)

legend_heatmap <- get_legend(
  p_heat1 + 
    theme(
      legend.position  = "bottom",
      legend.direction = "horizontal"
    ) +
    scale_fill_gradient2(
      low      = "#2166ac",
      mid      = "#f7f7f7", 
      high     = "#b2182b",
      midpoint = 0,
      limits   = c(-0.4, 0.4),
      breaks   = c(-0.4, -0.2, 0, 0.2, 0.4),
      labels   = c("-0.4", "-0.2", "0", "0.2", "0.4"),
      na.value = "grey90",
      oob      = scales::squish,
      name     = "Predation mortality rate change",
      guide    = guide_colorbar(
        title.position = "top",
        barwidth       = unit(10, "lines"),
        barheight      = unit(0.8, "lines"),
        ticks          = TRUE,
        ticks.colour   = "black",
        title.theme    = element_text(size = 15, color = "black"),  # title size
        label.theme    = element_text(size = 14, color = "black")   # numbers size
      )
    )
)
```

```{r}
p1_inset <- p1_inset + theme(plot.background = element_rect(fill = "white", colour = "black", linewidth = 1))
p2_inset <- p2_inset + theme(plot.background = element_rect(fill = "white", colour = "black", linewidth = 1))
# ── Titles ──
p1_titled <- p1 + ggtitle("With fishing") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        axis.text.y = element_text(size = 13))  # increase y-axis text size

p2_titled <- p2 + ggtitle("Without fishing") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
        axis.text.y = element_text(size = 13))

# ── Spectrums with insets ──
spectrum1 <- p1_titled + inset_element(p1_inset, left = 0.01, bottom = 0.6,
                                        right = 0.35, top = 0.99)
spectrum2 <- p2_titled + inset_element(p2_inset, left = 0.01, bottom = 0.6,
                                        right = 0.35, top = 0.99)

# ── Heatmap with larger predator labels ──
p_heat1 <- p_heat1 + theme(axis.text.y = element_text(size = 13))

p_heat2_clean <- p_heat2 +
  theme(axis.text.y  = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank())

# ── Final plots ──
final_plot1_clean <- (spectrum1 / p_heat1) +
  plot_layout(heights = c(4, 0.5)) &
  theme(legend.position = "none")

final_plot2_clean <- (spectrum2 / p_heat2_clean) +
  plot_layout(heights = c(4, 0.5)) &
  theme(legend.position = "none",
        axis.title.y    = element_blank())

library(cowplot)

# ── Legends ──
legend_species <- get_legend(
  p1 + theme(legend.position  = "bottom",
             legend.direction = "horizontal",
             legend.text      = element_text(size = 17),
             legend.title     = element_text(size = 16)) +
    guides(colour = guide_legend(nrow = 1,
                                 override.aes = list(linewidth = 2)))
)

legend_species <- get_legend(
  p1 + theme(legend.position  = "bottom",
             legend.direction = "horizontal",
             legend.text      = element_text(size = 17),
             legend.title     = element_text(size = 16)) +
    guides(colour = guide_legend(nrow = 1))
)

legend_heatmap <- get_legend(
  p_heat1 + theme(legend.position  = "bottom",
                  legend.direction = "horizontal"),
  title.theme    = element_text(size = 16, color = "black"),  # title size
        label.theme    = element_text(size = 15, color = "black")   # numbers size
)

# ── Combine ──
final_with_legend <- plot_grid(
  patchworkGrob(final_plot1_clean | final_plot2_clean),
  legend_heatmap,
  legend_species,
  ncol        = 1,
  rel_heights = c(1, 0.08, 0.10)
)

final_with_legend
# ggsave("final_with_legend0.tiff", final_with_legend, width = 14, height = 9, dpi = 600, bg = "white")

```

# biomass change in different stages of size spectrum

new w_max, w_mat.. size bins

```{r}
species  <- c("small_fish", "ruffe", "bream", "white_bream", "roach", "perch", "pikeperch")

# w mat25 new
targets <- c(0.7, 2, 203.887, 12.70513, 29.6453, 53.84554, 229.9023)
# w mat 50 new
targets <- c(1, 3, 303.108, 19.9652, 43.56044, 99.826, 376.3138)
# w mat75 new
targets <- c(1.7, 5, 506.9951, 32.67033, 73.20573, 153.6715, 606.2161)
# w max new
targets <- c(28, 144, 2783.0279, 474.9298, 719.9572, 1107.1611, 3630.0364)

data.frame(
  species = species,
  target  = targets,
  bin_index = sapply(targets, function(x) which.min(abs(w_bins - x))),
  bin_value = sapply(targets, function(x) w_bins[which.min(abs(w_bins - x))])
)

```

old w max bin

```{r}
# Get w_max for each species
w_max_species <- species_params(cur_model)$w_max
names(w_max_species) <- species_params(cur_model)$species

# Find bin index for each species w_max
data.frame(
  species   = species_params(cur_model)$species,
  w_max     = w_max_species,
  bin_index = sapply(w_max_species, function(x) which.min(abs(w_bins - x))),
  bin_value = sapply(w_max_species, function(x) w_bins[which.min(abs(w_bins - x))])
)
```

zero crossing bin

```{r}
zero_crossing <- nd_ratio %>%
  filter(!Legend %in% c("small_fish", "ruffe")) %>%
  group_by(Legend, Model) %>%
  arrange(w) %>%
  mutate(sign = sign(pct_change),
         sign_change = sign != lag(sign) & lag(sign) == 1 & sign == -1) %>%
  filter(sign_change) %>%
  slice(n()) %>%  # take LAST crossing
  summarise(
    w_zero_crossing = w,
    bin_index = which.min(abs(w_bins - w)),
    .groups = "drop"
  )

print(zero_crossing)
```

new w mat size bin

```{r}
data.frame(
  species   = species_params(cur_model)$species,
  w_mat_new = w_max_t50$w_mat_t50,
  bin_index = sapply(w_max_t50$w_mat_t50, function(x) which.min(abs(w_bins - x))),
  bin_value = sapply(w_max_t50$w_mat_t50, function(x) w_bins[which.min(abs(w_bins - x))])
)
```

new w mat25 size bin

```{r}
data.frame(
  species   = species_params(cur_model)$species,
  w_mat_new = w_max_t50$w_mat_t25,
  bin_index = sapply(w_max_t50$w_mat_t25, function(x) which.min(abs(w_bins - x))),
  bin_value = sapply(w_max_t50$w_mat_t25, function(x) w_bins[which.min(abs(w_bins - x))])
)
```

what percentage of the total initial biomass was lost/gained specifically in the size classes?

Absolute biomass change in each size class as a proportion of total initial biomass. This naturally weights changes by the initial biomass concentration --- a 87% loss in a size class holding only 1% of total biomass is ecologically less significant than a 12% loss in a size class holding 60% of total biomass. Size classes that were both biomass-rich and strongly affected contribute most to the total change, making this a more informative measure than within-class percentage change alone.

```{r}
n_firstM <- all_model_results$M7_kappa9_lambda2.05extMort01_0.333.rds$species_3_4_5_6_7
n <- N(n_firstM)
w <- w(n_firstM@params)

# Total initial biomass for bream
bins_all <- 1:200
B_total1.3 <- sum(n[1, 3, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])

# Each class contribution relative to TOTAL initial biomass
bins_big  <- 191:200
bins_mid  <- 188:190
bins_mat  <- 162:187
bins_small <- 1:161

pct_big   <- 100 * (sum(n[1,3,bins_big]   * w_bins[bins_big]   * dw(n_firstM@params)[bins_big])   - 
                    sum(n[51,3,bins_big]   * w_bins[bins_big]   * dw(n_firstM@params)[bins_big]))   / B_total1.3

pct_mid   <- 100 * (sum(n[1,3,bins_mid]   * w_bins[bins_mid]   * dw(n_firstM@params)[bins_mid])   - 
                    sum(n[51,3,bins_mid]   * w_bins[bins_mid]   * dw(n_firstM@params)[bins_mid]))   / B_total1.3

pct_mat   <- 100 * (sum(n[1,3,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat])   - 
                    sum(n[51,3,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat]))   / B_total1.3

pct_small <- 100 * (sum(n[1,3,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small]) - 
                    sum(n[51,3,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small])) / B_total1.3

# These will now sum to the total % change
cat("big:  ", pct_big, "\n")
cat("mid:  ", pct_mid, "\n")
cat("mat:  ", pct_mat, "\n")
cat("small:", pct_small, "\n")
cat("total:", pct_big + pct_mid + pct_mat + pct_small, "\n")

# Verify against direct total
B_total51.3 <- sum(n[51, 3, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])
cat("direct total % change:", 100 * (B_total1.3 - B_total51.3) / B_total1.3, "\n")
```

white bream

```{r}
# Total initial biomass for white bream
bins_all <- 1:200
B_total1.4 <- sum(n[1, 4, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])

# Each class contribution relative to TOTAL initial biomass
bins_big  <- 169:200
bins_mat  <- 127:168
bins_small <- 1:126

pct_big   <- 100 * (sum(n[1,4,bins_big]   * w_bins[bins_big]   * dw(n_firstM@params)[bins_big])   - 
                    sum(n[51,4,bins_big]   * w_bins[bins_big]   * dw(n_firstM@params)[bins_big]))   / B_total1.4

pct_mat   <- 100 * (sum(n[1,4,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat])   - 
                    sum(n[51,4,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat]))   / B_total1.4

pct_small <- 100 * (sum(n[1,4,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small]) - 
                    sum(n[51,4,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small])) / B_total1.4

# These will now sum to the total % change
cat("big:  ", pct_big, "\n")
cat("mat:  ", pct_mat, "\n")
cat("small:", pct_small, "\n")
cat("total:", pct_big + pct_mat + pct_small, "\n")

# Verify against direct total
B_total51.4 <- sum(n[51, 4, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])
cat("direct total % change:", 100 * (B_total1.4 - B_total51.4) / B_total1.4, "\n")
```

```{r}
# Total initial biomass for roach
bins_all <- 1:200
B_total1.5 <- sum(n[1, 5, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])

# Each class contribution relative to TOTAL initial biomass
bins_big  <- 174:200
bins_mid  <- 157:173
bins_mat  <- 137:156
bins_small <- 1:136

pct_big   <- 100 * (sum(n[1,5,bins_big]   * w_bins[bins_big]   * dw(n_firstM@params)[bins_big])   - 
                    sum(n[51,5,bins_big]   * w_bins[bins_big]   * dw(n_firstM@params)[bins_big]))   / B_total1.5

pct_mid   <- 100 * (sum(n[1,5,bins_mid]   * w_bins[bins_mid]   * dw(n_firstM@params)[bins_mid])   - 
                    sum(n[51,5,bins_mid]   * w_bins[bins_mid]   * dw(n_firstM@params)[bins_mid]))   / B_total1.5

pct_mat   <- 100 * (sum(n[1,5,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat])   - 
                    sum(n[51,5,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat]))   / B_total1.5

pct_small <- 100 * (sum(n[1,5,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small]) - 
                    sum(n[51,5,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small])) / B_total1.5

# These will now sum to the total % change
cat("big:  ", pct_big, "\n")
cat("mid:  ", pct_mid, "\n")
cat("mat:  ", pct_mat, "\n")
cat("small:", pct_small, "\n")
cat("total:", pct_big + pct_mid + pct_mat + pct_small, "\n")

# Verify against direct total
B_total51.5 <- sum(n[51, 5, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])
cat("direct total % change:", 100 * (B_total1.5 - B_total51.5) / B_total1.5, "\n")
```

```{r}
# Total initial biomass for perch
bins_all <- 1:200
B_total1.6 <- sum(n[1, 6, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])

# Each class contribution relative to TOTAL initial biomass
bins_big  <- 179:200
bins_mid  <- 172:178
bins_mat  <- 148:171
bins_small <- 1:147

pct_big   <- 100 * (sum(n[1,6,bins_big]   * w_bins[bins_big]   * dw(n_firstM@params)[bins_big])   - 
                    sum(n[51,6,bins_big]   * w_bins[bins_big]   * dw(n_firstM@params)[bins_big]))   / B_total1.6

pct_mid   <- 100 * (sum(n[1,6,bins_mid]   * w_bins[bins_mid]   * dw(n_firstM@params)[bins_mid])   - 
                    sum(n[51,6,bins_mid]   * w_bins[bins_mid]   * dw(n_firstM@params)[bins_mid]))   / B_total1.6

pct_mat   <- 100 * (sum(n[1,6,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat])   - 
                    sum(n[51,6,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat]))   / B_total1.6

pct_small <- 100 * (sum(n[1,6,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small]) - 
                    sum(n[51,6,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small])) / B_total1.6

# These will now sum to the total % change
cat("big:  ", pct_big, "\n")
cat("mid:  ", pct_mid, "\n")
cat("mat:  ", pct_mat, "\n")
cat("small:", pct_small, "\n")
cat("total:", pct_big + pct_mid + pct_mat + pct_small, "\n")

# Verify against direct total
B_total51.6 <- sum(n[51, 6, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])
cat("direct total % change:", 100 * (B_total1.6 - B_total51.6) / B_total1.6, "\n")

```

```{r}
# Total initial biomass for pikeperch
bins_all <- 1:200
B_total1.7 <- sum(n[1, 7, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])

# Each class contribution relative to TOTAL initial biomass
bins_big  <- 195:200
bins_mid  <- 193:194
bins_mat  <- 165:192
bins_small <- 1:164

pct_big   <- 100 * (sum(n[1,7,bins_big]   * w_bins[bins_big]   * dw(n_firstM@params)[bins_big])   - 
                    sum(n[51,7,bins_big]   * w_bins[bins_big]   * dw(n_firstM@params)[bins_big]))   / B_total1.7

pct_mid   <- 100 * (sum(n[1,7,bins_mid]   * w_bins[bins_mid]   * dw(n_firstM@params)[bins_mid])   - 
                    sum(n[51,7,bins_mid]   * w_bins[bins_mid]   * dw(n_firstM@params)[bins_mid]))   / B_total1.7

pct_mat   <- 100 * (sum(n[1,7,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat])   - 
                    sum(n[51,7,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat]))   / B_total1.7

pct_small <- 100 * (sum(n[1,7,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small]) - 
                    sum(n[51,7,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small])) / B_total1.7

# These will now sum to the total % change
cat("big:  ", pct_big, "\n")
cat("mid:  ", pct_mid, "\n")
cat("mat:  ", pct_mat, "\n")
cat("small:", pct_small, "\n")
cat("total:", pct_big + pct_mid + pct_mat + pct_small, "\n")

# Verify against direct total
B_total51.7 <- sum(n[51, 7, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])
cat("direct total % change:", 100 * (B_total1.7 - B_total51.7) / B_total1.7, "\n")
```

```{r}
# Total initial biomass for ruffe
bins_all <- 1:200
B_total1.2 <- sum(n[1, 2, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])

# Each class contribution relative to TOTAL initial biomass
bins_mat  <- 103:200
bins_small <- 1:102

pct_mat   <- 100 * (sum(n[1,2,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat])   - 
                    sum(n[51,2,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat]))   / B_total1.2

pct_small <- 100 * (sum(n[1,2,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small]) - 
                    sum(n[51,2,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small])) / B_total1.2

# These will now sum to the total % change
cat("mat:  ", pct_mat, "\n")
cat("small:", pct_small, "\n")
cat("total:", pct_mat + pct_small, "\n")

# Verify against direct total
B_total51.2 <- sum(n[51, 2, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])
cat("direct total % change:", 100 * (B_total1.2 - B_total51.2) / B_total1.2, "\n")
```

```{r}
# Total initial biomass for small fish
bins_all <- 1:200
B_total1.1 <- sum(n[1, 1, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])

# Each class contribution relative to TOTAL initial biomass

bins_mat  <- 89:200
bins_small <- 1:88

pct_mat   <- 100 * (sum(n[1,1,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat])   - 
                    sum(n[51,1,bins_mat]   * w_bins[bins_mat]   * dw(n_firstM@params)[bins_mat]))   / B_total1.1

pct_small <- 100 * (sum(n[1,1,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small]) - 
                    sum(n[51,1,bins_small] * w_bins[bins_small] * dw(n_firstM@params)[bins_small])) / B_total1.1

# These will now sum to the total % change
cat("mat:  ", pct_mat, "\n")
cat("small:", pct_small, "\n")
cat("total:", pct_mat + pct_small, "\n")

# Verify against direct total
B_total51.1 <- sum(n[51, 1, bins_all] * w_bins[bins_all] * dw(n_firstM@params)[bins_all])
cat("direct total % change:", 100 * (B_total1.1 - B_total51.1) / B_total1.1, "\n")
```

size change table test needed for below code

```{r}
n_firstM <- all_model_results$M5_kappa7_lambda2.05extMort01_0.333.rds$species_3_4_5_6_7
n <- N(n_firstM)
w <- w(n_firstM@params)

#install.packages("gt")
library(gt)

# ── helper ────────────────────────────────────────────────────────────────────
compute_biomass_pct <- function(sp_idx, size_classes) {
  # size_classes: named list of bin ranges, e.g. list(small=1:88, mat=89:200)
  bins_all <- 1:200
  dw_vec   <- dw(n_firstM@params)
  w        <- w_bins

  B0 <- sum(n[1,  sp_idx, bins_all] * w[bins_all] * dw_vec[bins_all])
  B1 <- sum(n[51, sp_idx, bins_all] * w[bins_all] * dw_vec[bins_all])

 # In compute_biomass_pct, change the sign:
pcts <- sapply(size_classes, function(bins) {
  100 * (
    sum(n[51, sp_idx, bins] * w[bins] * dw_vec[bins]) -   # B1 first
    sum(n[1,  sp_idx, bins] * w[bins] * dw_vec[bins])
  ) / B0
})

c(pcts, total = sum(pcts), direct = 100 * (B1 - B0) / B0)
}

# ── species definitions ───────────────────────────────────────────────────────
species_list <- list(
  list(name = "Small fish",   idx = 1, classes = list(small = 1:85,   mat = 86:96, mid = 97:132, big = 133:200)),
  list(name = "Ruffe",        idx = 2, classes = list(small = 1:98,  mat = 99:110, mid = 111:152, big = 153:200)),
  list(name = "Bream",        idx = 3, classes = list(small = 1:157,  mat = 158:168, mid = 169:190, big = 191:200)),
  list(name = "White bream",  idx = 4, classes = list(small = 1:121,  mat = 122:134, mid = 135:168, big = 169:200)),
  list(name = "Roach",        idx = 5, classes = list(small = 1:132,  mat = 133:144, mid = 145:173, big = 174:200)),
  list(name = "Perch",        idx = 6, classes = list(small = 1:140,  mat = 141:153, mid = 154:178, big = 179:200)),
  list(name = "Pikeperch",    idx = 7, classes = list(small = 1:158,  mat = 159:171, mid = 172:194, big = 195:200))
)

```

```{r}
# ── compute for all models ────────────────────────────────────────────────────
all_classes <- c("small", "mat", "mid", "big")

compute_biomass_pct_model <- function(model_results, species_list) {
  n_sim  <- model_results$species_3_4_5_6_7
  n      <- N(n_sim)
  dw_vec <- dw(n_sim@params)
  w_bins <- w(n_sim@params)

  lapply(species_list, function(sp) {
    bins_all <- 1:200
    B0 <- sum(n[1,  sp$idx, bins_all] * w_bins[bins_all] * dw_vec[bins_all])
    B1 <- sum(n[51, sp$idx, bins_all] * w_bins[bins_all] * dw_vec[bins_all])

    pcts <- sapply(sp$classes, function(bins) {
      100 * (
        sum(n[51, sp$idx, bins] * w_bins[bins] * dw_vec[bins]) -
        sum(n[1,  sp$idx, bins] * w_bins[bins] * dw_vec[bins])
      ) / B0
    })

    row        <- as.list(pcts[all_classes])
    row$total  <- 100 * (B1 - B0) / B0
    row$species <- sp$name
    as.data.frame(row, check.names = FALSE)
  }) |> bind_rows()
}

# Run across all 5 models
all_results <- lapply(names(all_model_results), function(m) {
  compute_biomass_pct_model(all_model_results[[m]], species_list) |>
    mutate(model = m)
}) |> bind_rows()

# ── summarise: min–max range per species × size class ────────────────────────
range_df <- all_results |>
  group_by(species) |>
  summarise(across(c(small, mat, mid, big, total),
                   list(min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE))),
            .groups = "drop")

# ── format as "min – max" labels ─────────────────────────────────────────────
format_range <- function(min_val, max_val) {
  min_r <- round(min_val, 0)
  max_r <- round(max_val, 0)
  ifelse(
    is.na(min_val), NA_character_,
    ifelse(
      min_r == max_r,
      as.character(min_r),
      paste0(min_r, " – ", max_r)
    )
  )
}

results_displayF <- range_df |>
  mutate(
    small = format_range(small_min, small_max),
    mat   = format_range(mat_min,   mat_max),
    mid   = format_range(mid_min,   mid_max),
    big   = format_range(big_min,   big_max),
    total = format_range(total_min, total_max)
  ) |>
  select(species, small, mat, mid, big, total)

# ── total row: sum across species per model first, then range ─────────────────
totals_row <- all_results |>
  group_by(model) |>
  summarise(across(c(small, mat, mid, big, total),
                   ~sum(., na.rm = TRUE)),
            .groups = "drop") |>
  summarise(across(c(small, mat, mid, big, total),
                   list(min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE)))) |>
  mutate(
    small   = format_range(small_min, small_max),
    mat     = format_range(mat_min,   mat_max),
    mid     = format_range(mid_min,   mid_max),
    big     = format_range(big_min,   big_max),
    total   = format_range(total_min, total_max),
    species = "Total"
  ) |>
  select(species, small, mat, mid, big, total)

results_displayF <- bind_rows(results_displayF, totals_row)

# ── gt table ──────────────────────────────────────────────────────────────────
results_displayF |>
  gt(rowname_col = "species") |>
  tab_header(
    title    = "Biomass change by size class",
    subtitle = "Range (min – max) across 5 models, % relative to initial total biomass at year 50"
  ) |>
  cols_label(
    small = "Small",
    mat   = "Mature",
    mid   = "Mid",
    big   = "Large",
    total = "Total"
  ) |>
  sub_missing(missing_text = "—") |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#f0f0f0"),
    locations = cells_body(columns = total)
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_body(columns = total)
  ) |>
  tab_style(
    style     = cell_text(weight = "bold", style = "italic"),
    locations = cells_stub(rows = "Total")
  ) |>
  tab_style(
    style     = cell_fill(color = "#e8e8e8"),
    locations = cells_body(rows = species == "Total")
  ) |>
  tab_spanner(
    label   = "Size class contribution (% of B₀)",
    columns = c(small, mat, mid, big)
  ) |>
  opt_stylize(style = 6, color = "blue")

```

```{r}
# Sum of size class totals row (small + mat + mid + big summed across species)
all_results |>
  group_by(model) |>
  summarise(size_sum = sum(small, na.rm = TRUE) + 
                       sum(mat,   na.rm = TRUE) + 
                       sum(mid,   na.rm = TRUE) + 
                       sum(big,   na.rm = TRUE)) |>
  summarise(
    min_size = min(size_sum),
    max_size = max(size_sum)
  )

# Sum of total column (direct B1-B0 calculation summed across species)
all_results |>
  group_by(model) |>
  summarise(total_sum = sum(total, na.rm = TRUE)) |>
  summarise(
    min_total = min(total_sum),
    max_total = max(total_sum)
  )
```

# no fishing number change

```{r}
n_firstM <- all_model_results$M5_kappa7_lambda2.05extMort01_0.333.rds$species_3_4_5_6_7
n <- N(n_firstM)
w <- w(n_firstM@params)

#install.packages("gt")
library(gt)

# ── helper ────────────────────────────────────────────────────────────────────
compute_biomass_pct <- function(sp_idx, size_classes) {
  # size_classes: named list of bin ranges, e.g. list(small=1:88, mat=89:200)
  bins_all <- 1:200
  dw_vec   <- dw(n_firstM@params)
  w        <- w_bins

  B0 <- sum(n[1,  sp_idx, bins_all] * w[bins_all] * dw_vec[bins_all])
  B1 <- sum(n[51, sp_idx, bins_all] * w[bins_all] * dw_vec[bins_all])

 # In compute_biomass_pct, change the sign:
pcts <- sapply(size_classes, function(bins) {
  100 * (
    sum(n[51, sp_idx, bins] * w[bins] * dw_vec[bins]) -   # B1 first
    sum(n[1,  sp_idx, bins] * w[bins] * dw_vec[bins])
  ) / B0
})

c(pcts, total = sum(pcts), direct = 100 * (B1 - B0) / B0)
}

# ── species definitions ───────────────────────────────────────────────────────
species_list <- list(
  list(name = "Small fish",   idx = 1, classes = list(small = 1:88,   mat = 89:200)),
  list(name = "Ruffe",        idx = 2, classes = list(small = 1:102,  mat = 103:200)),
  list(name = "Bream",        idx = 3, classes = list(small = 1:161,  mat = 162:190, big = 191:200)),
  list(name = "White bream",  idx = 4, classes = list(small = 1:126,  mat = 127:168, big = 169:200)),
  list(name = "Roach",        idx = 5, classes = list(small = 1:136,  mat = 137:173, big = 174:200)),
  list(name = "Perch",        idx = 6, classes = list(small = 1:147,  mat = 148:178, big = 179:200)),
  list(name = "Pikeperch",    idx = 7, classes = list(small = 1:164,  mat = 165:194, big = 195:200))
)

```

```{r}
# ── compute for all models ────────────────────────────────────────────────────
all_classes <- c("small", "mat", "big")

compute_biomass_pct_model <- function(model_results, species_list) {
  n_sim  <- model_results$species_3_4_5_6_7
  n      <- N(n_sim)
  dw_vec <- dw(n_sim@params)
  w_bins <- w(n_sim@params)

  lapply(species_list, function(sp) {
    bins_all <- 1:200
    B0 <- sum(n[1,  sp$idx, bins_all] * w_bins[bins_all] * dw_vec[bins_all])
    B1 <- sum(n[51, sp$idx, bins_all] * w_bins[bins_all] * dw_vec[bins_all])

    pcts <- sapply(sp$classes, function(bins) {
      100 * (
        sum(n[51, sp$idx, bins] * w_bins[bins] * dw_vec[bins]) -
        sum(n[1,  sp$idx, bins] * w_bins[bins] * dw_vec[bins])
      ) / B0
    })

    row        <- as.list(pcts[all_classes])
    row$total  <- 100 * (B1 - B0) / B0
    row$species <- sp$name
    as.data.frame(row, check.names = FALSE)
  }) |> bind_rows()
}

# Run across all 5 models
all_results <- lapply(names(all_model_results), function(m) {
  compute_biomass_pct_model(all_model_results[[m]], species_list) |>
    mutate(model = m)
}) |> bind_rows()

# ── summarise: min–max range per species × size class ────────────────────────
range_df <- all_results |>
  group_by(species) |>
  summarise(across(c(small, mat, big, total),
                   list(min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE))),
            .groups = "drop")

# ── format as "min – max" labels ─────────────────────────────────────────────
format_range <- function(min_val, max_val) {
  min_r <- round(min_val, 0)
  max_r <- round(max_val, 0)
  ifelse(
    is.na(min_val), NA_character_,
    ifelse(
      min_r == max_r,
      as.character(min_r),
      paste0(min_r, " – ", max_r)
    )
  )
}

results_display <- range_df |>
  mutate(
    small = format_range(small_min, small_max),
    mat   = format_range(mat_min,   mat_max),
    big   = format_range(big_min,   big_max),
    total = format_range(total_min, total_max)
  ) |>
  select(species, small, mat, big, total)

# ── total row: sum across species per model first, then range ─────────────────
totals_row <- all_results |>
  group_by(model) |>
  summarise(across(c(small, mat, big, total),
                   ~sum(., na.rm = TRUE)),
            .groups = "drop") |>
  summarise(across(c(small, mat, big, total),
                   list(min = ~min(., na.rm = TRUE),
                        max = ~max(., na.rm = TRUE)))) |>
  mutate(
    small   = format_range(small_min, small_max),
    mat     = format_range(mat_min,   mat_max),
    big     = format_range(big_min,   big_max),
    total   = format_range(total_min, total_max),
    species = "Total"
  ) |>
  select(species, small, mat, big, total)

results_display <- bind_rows(results_display, totals_row)

# ── gt table ──────────────────────────────────────────────────────────────────
results_display |>
  gt(rowname_col = "species") |>
  tab_header(
    title    = "Biomass change by size class",
    subtitle = "Range (min – max) across 5 models, % relative to initial total biomass at year 50"
  ) |>
  cols_label(
    small = "Small",
    mat   = "Mature",
    big   = "Large",
    total = "Total"
  ) |>
  sub_missing(missing_text = "—") |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style     = cell_fill(color = "#f0f0f0"),
    locations = cells_body(columns = total)
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_body(columns = total)
  ) |>
  tab_style(
    style     = cell_text(weight = "bold", style = "italic"),
    locations = cells_stub(rows = "Total")
  ) |>
  tab_style(
    style     = cell_fill(color = "#e8e8e8"),
    locations = cells_body(rows = species == "Total")
  ) |>
  tab_spanner(
    label   = "Size class contribution (% of B₀)",
    columns = c(small, mat, big)
  ) |>
  opt_stylize(style = 6, color = "blue")

```

```{r}
# Sum of size class totals row (small + mat + mid + big summed across species)
all_results |>
  group_by(model) |>
  summarise(size_sum = sum(small, na.rm = TRUE) + 
                       sum(mat,   na.rm = TRUE) + 
                       sum(big,   na.rm = TRUE)) |>
  summarise(
    min_size = min(size_sum),
    max_size = max(size_sum)
  )

# Sum of total column (direct B1-B0 calculation summed across species)
all_results |>
  group_by(model) |>
  summarise(total_sum = sum(total, na.rm = TRUE)) |>
  summarise(
    min_total = min(total_sum),
    max_total = max(total_sum)
  )
```

# Pred kernel

```{r}
cur_model <- read_rds("M5_kappa7_lambda2.05extMort01_0.333.rds")
pk     <- getPredKernel(cur_model)
w_prey <- cur_model@w_full
w_grid <- cur_model@w

predators <- list(
  list(species = "pikeperch", weight = 3000),
  list(species = "perch",     weight = 800)
)

pk_df <- purrr::map_dfr(predators, function(pred) {
  sp_idx   <- which(dimnames(pk)[[1]] == pred$species)
  w_idx    <- which.min(abs(w_grid - pred$weight))
  actual_w <- w_grid[w_idx]
  phi      <- pk[sp_idx, w_idx, ]
  
  data.frame(
    prey_w   = w_prey,
    phi      = phi,
    species  = pred$species,
    actual_w = actual_w,
    peak_w   = w_prey[which.max(phi)],  # prey weight at peak kernel
    label    = paste0(pred$species, " (", pred$weight, " g)")
  )
})

ggplot(pk_df, aes(x = prey_w, y = phi)) +
  geom_line(linewidth = 1.1, color = "grey20") +
  geom_vline(aes(xintercept = actual_w), linetype = "dashed",
             color = "gray40", linewidth = 0.5) +
  geom_vline(aes(xintercept = peak_w), linetype = "dashed",
             color = "grey50", linewidth = 0.5) +
  geom_text(aes(x = peak_w, y = max(phi) * 0.6,
                label = paste0("Peak: ", round(peak_w, 1), " g")),
            hjust = -0.1, size = 6, color = "grey40",
            data = pk_df %>% distinct(label, peak_w, .keep_all = TRUE)) +
  coord_cartesian(xlim = c(0, 500)) +
  scale_x_continuous(labels = scales::comma) +
  facet_wrap2(~ label, axes = "all") +   # forces axes on all facets
  labs(
    x = "Prey weight (g)",
    y = "Predation kernel φ"
  ) +
  My_theme +
  theme(
    axis.line.x.bottom = element_line(color = "black", linewidth = 0.5),
    axis.line.y.left   = element_line(color = "black", linewidth = 0.5)
  )
```

Predation change

predation rate = integral over all predator sizes of (predation kernel × predator abundance)
```{r}
sim    <- all_model_results[["M5_kappa7_lambda2.05extMort01_0.333.rds"]][["species_3_4_5_6_7"]]
params <- sim@params

t_first <- 1
t_last  <- dim(sim@n)[1]

n_first    <- sim@n[t_first, , ]
n_last     <- sim@n[t_last, , ]
n_pp_first <- sim@n_pp[t_first, ]
n_pp_last  <- sim@n_pp[t_last, ]

# Predation rate over prey size grid (w_full)
pr_first <- getPredRate(params, n = n_first, n_pp = n_pp_first)
pr_last  <- getPredRate(params, n = n_last,  n_pp = n_pp_last)

sp_idx   <- which(dimnames(pr_first)[[1]] == "pikeperch")
w_prey   <- params@w_full

pr_df <- bind_rows(
  data.frame(prey_w = w_prey, pred_rate = pr_first[sp_idx, ], version = "First time step (t=0)"),
  data.frame(prey_w = w_prey, pred_rate = pr_last[sp_idx, ],  version = "Last time step (t=50)")
)

ggplot(pr_df, aes(x = prey_w, y = pred_rate, color = version)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("First time step (t=0)" = "#93AA00",
                                "Last time step (t=50)" = "#2166ac")) +
  coord_cartesian(xlim = c(0, 500)) +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title    = "Predation rate — pikeperch",
    subtitle = "Scenario: w_mat decline in species 3–7",
    x        = "Prey weight (g)",
    y        = "Predation rate",
    color    = NULL,
    caption  = "Model: M5_kappa7_lambda2.05extMort01_0.333"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.line          = element_line(color = "black", linewidth = 0.5),
    axis.text          = element_text(size = 9, color = "black"),
    axis.title         = element_text(size = 11, face = "bold"),
    axis.ticks         = element_line(color = "black", linewidth = 0.5),
    plot.title         = element_text(size = 13, face = "bold"),
    plot.subtitle      = element_text(size = 10, color = "gray40"),
    plot.background    = element_rect(fill = "white", color = NA),
    plot.margin        = margin(10, 15, 10, 10),
    legend.position    = "top"
  )

```


# Yields with true effort

```{r}
psi_at_t <- function(params, t) {
  op  <- params@other_params$psi
  idx <- floor(t) + 1
  idx <- max(1, min(idx, dim(op)[1]))   # clamp to valid range
  list(psi = op[idx, , ])
}

# === REALISTIC YEAR-BY-YEAR EFFORT: baseline vs decline ===

# --- 0. Effort matrix + spin-up state (unchanged) ---
build_realistic_effort <- function(params) {
  current_gear        <- gear_params(params)
  current_gear$gear   <- current_gear$species
  gear_params(params) <- current_gear
  params <- setParams(params)

  f_main_complete <- cbind(small_fish = rep(0, nrow(f_main)),
                           ruffe      = rep(0, nrow(f_main)),
                           f_main)
  rownames(f_main_complete) <- as.character(years_data)

  sim2 <- project(params, effort = initial_effort, t_max = 300, dt = 0.1)

  species_order   <- species_params(sim2@params)$species
  f_main_ordered  <- f_main_complete[, species_order]
  relative_effort <- sweep(f_main_ordered, 2,
                           gear_params(sim2@params)$catchability, "/")
  relative_effort[is.nan(relative_effort)] <- 0

  list(
    effort_full = add_ramp_up(relative_effort),   # rows named 1893:2023
    n_init      = sim2@n[300, , ],
    npp_init    = sim2@n_pp[300, ]
  )
}

# --- 1. psi-array builder (unchanged; 31 rows for times 0:30) ---
make_psi_array <- function(params, decrease_rate, n_years = 30) {
  times   <- 0:n_years
  species <- species_params(params)$species
  n_sp    <- length(species)
  w       <- params@w

  psi_array <- array(NA, dim = c(length(times), n_sp, length(w)),
                     dimnames = list(time = times, sp = species, w = w))

  w_mat_array   <- matrix(NA, length(times), n_sp)
  w_mat25_array <- w_mat_array
  w_max_array   <- w_mat_array

  w_mat_cur   <- params@species_params$w_mat
  w_mat25_cur <- params@species_params$w_mat25
  w_max_cur   <- params@species_params$w_max

  for (i in seq_along(times)) {
    w_mat_array[i, ]   <- w_mat_cur
    w_mat25_array[i, ] <- w_mat25_cur
    w_max_array[i, ]   <- w_max_cur
    w_mat_cur   <- w_mat_cur   * (1 - decrease_rate)
    w_mat25_cur <- w_mat25_cur * (1 - decrease_rate)
    w_max_cur   <- w_max_cur   * (1 - decrease_rate)
  }

  m <- params@species_params$m
  n <- params@species_params$n

  for (t in seq_along(times)) {
    for (sp in seq_len(n_sp)) {
      w_mat_t   <- w_mat_array[t, sp]
      w_mat25_t <- w_mat25_array[t, sp]
      w_rmax    <- w_max_array[t, sp]
      if (!is.na(w_mat_t) && !is.na(w_mat25_t) && w_mat_t > w_mat25_t) {
        U        <- log(3) / log(w_mat_t / w_mat25_t)
        maturity <- (1 + (w / w_mat_t)^(-U))^(-1)
        maturity[maturity < 1e-08] <- 0
        repro_prop <- (w / w_rmax)^(m[sp] - n[sp])
        repro_prop[repro_prop > 1] <- 1
        psi <- maturity * repro_prop
        psi[psi > 1] <- 1
        psi[w > w_rmax] <- 1
        psi_array[t, sp, ] <- psi
      } else {
        psi_array[t, sp, ] <- 0
      }
    }
  }
  psi_array
}

# --- 2. Run one model: spin up to 1992, then project 1993-2022 ---
#     under YEAR-BY-YEAR historical effort + time-varying psi.
decline_start <- 1993
decline_end   <- 2022                          # 30 steps -> times 0:30
n_years       <- decline_end - decline_start + 1   # = 30

run_realistic_decline <- function(params, decrease_rate) {
  re   <- build_realistic_effort(params)
  eff  <- re$effort_full
  yrs  <- as.numeric(rownames(eff))

  # State at end of 1992 (start of decline window)
  hist_rows <- yrs <= (decline_start - 1)       # 1893..1992
  sim_hist  <- project(params, effort = eff[hist_rows, , drop = FALSE],
                        dt = 0.1, t_save = 1,
                        initial_n   = re$n_init,
                        initial_n_pp = re$npp_init)
  last <- dim(sim_hist@n)[1]
  n0   <- sim_hist@n[last, , ]
  npp0 <- sim_hist@n_pp[last, ]

  # Projection-window effort: 1993..2022 (year-by-year, real variation)
  proj_rows   <- yrs >= decline_start & yrs <= decline_end
  effort_proj <- eff[proj_rows, , drop = FALSE]   # 30 rows

  p2 <- params
  p2@other_params$psi <- make_psi_array(p2, decrease_rate, n_years)
  p2 <- setRateFunction(p2, "ERepro", "timeVaryingERepro")

  project(p2, dt = 0.1, t_save = 1,
          effort      = effort_proj,
          initial_n   = n0,
          initial_n_pp = npp0)
}

# --- 3. Loop models x {baseline, decline} ---
decline_rate  <- c(0, 0, 0.01, 0.01, 0.01, 0.01, 0.01)
baseline_rate <- rep(0, 7)

realistic_results <- list()
for (mn in names(models)) {
  cat("Model:", mn, "\n")
  realistic_results[[mn]] <- list(
    Baseline                = run_realistic_decline(models[[mn]], baseline_rate),
    `With maturity decline` = run_realistic_decline(models[[mn]], decline_rate)
  )
}

# --- 4. Tidy yields ---
get_yield_df <- function(res, mn, scen) {
  yld <- getYield(res)                 # rows = saved timesteps
  n_t <- nrow(yld)
  yld %>% melt() %>%
    mutate(model = mn, scenario = scen) %>%
    filter(!sp %in% c("small_fish", "ruffe")) %>%
    # overwrite whatever 'time' melt produced with a clean step index
    group_by(sp) %>%
    mutate(step = row_number() - 1) %>%   # 0,1,2,... per species
    ungroup() %>%
    mutate(year = decline_start + step)   # 1993, 1994, ...
}

y_realistic <- bind_rows(lapply(names(realistic_results), function(mn) {
  bind_rows(
    get_yield_df(realistic_results[[mn]]$Baseline, mn, "Baseline"),
    get_yield_df(realistic_results[[mn]]$`With maturity decline`, mn, "With maturity decline")
  )
}))

```

```{r}
combined_biomass_data <- data.frame()

for (model_name in names(realistic_results)) {
  for (scenario_name in names(realistic_results[[model_name]])) {
    sim <- realistic_results[[model_name]][[scenario_name]]
    biomass <- getBiomass(sim)

    biomass_long <- as.data.frame(biomass) %>%
      mutate(time = as.numeric(rownames(biomass))) %>%
      pivot_longer(cols = -time, names_to = "species", values_to = "biomass") %>%
      mutate(model = model_name, scenario = scenario_name)

    combined_biomass_data <- bind_rows(combined_biomass_data, biomass_long)
  }
}

biomass_change_summary <- combined_biomass_data %>%
  group_by(model, scenario, species) %>%
  mutate(biomass_change_percent = (biomass - first(biomass)) / first(biomass) * 100) %>%
  ungroup() %>%
  group_by(scenario, species, time) %>%
  summarise(
    mean_change   = mean(biomass_change_percent),
    median_change = median(biomass_change_percent),
    min_change    = min(biomass_change_percent),
    max_change    = max(biomass_change_percent),
    .groups = 'drop'
  ) %>%
  mutate(scenario_label = scenario)   # "Baseline" / "With maturity decline"

species_colors <- c(
  "small_fish"  = "#FCC300",
  "ruffe"       = "#44AA99",
  "bream"       = "#b2182b",
  "white_bream" = "#661100",
  "roach"       = "#AA4499",
  "perch"       = "#2166ac",
  "pikeperch"   = "#93AA00"
)

p0 <- ggplot(biomass_change_summary,
             aes(x = time, y = mean_change, color = species, fill = species)) +
  geom_ribbon(aes(ymin = min_change, ymax = max_change), alpha = 0.2, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  facet_wrap2(~ scenario_label, scales = "fixed", ncol = 1,
              axes = "all", remove_labels = "x") +
  labs(x = "Time (years)", y = "Biomass Change (%)",
       color = "Species", fill = "Species") +
  scale_color_manual(values = species_colors) +
  scale_fill_manual(values = species_colors) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    panel.border       = element_blank(),
    axis.line.x.bottom = element_line(color = "black", linewidth = 0.5),
    axis.line.y.left   = element_line(color = "black", linewidth = 0.5),
    strip.background   = element_blank(),
    strip.text  = element_text(size = 10, face = "bold", hjust = 0, margin = margin(b = 5)),
    axis.line   = element_line(color = "black", linewidth = 0.5),
    axis.text   = element_text(size = 9, color = "black"),
    axis.title  = element_text(size = 11, face = "bold"),
    axis.ticks  = element_line(color = "black", linewidth = 0.5),
    legend.position   = "right",
    legend.title      = element_text(size = 10, face = "bold"),
    legend.text       = element_text(size = 9),
    legend.key        = element_blank(),
    legend.background = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin     = margin(10, 10, 10, 10),
    plot.subtitle   = element_text(size = 9, hjust = 0.5)
  )

print(p0)

baseline_diff_summary <- combined_biomass_data %>%
  group_by(model, scenario, species) %>%
  mutate(biomass_change_percent = (biomass - first(biomass)) / first(biomass) * 100) %>%
  ungroup() %>%
  select(model, species, time, scenario, biomass_change_percent) %>%
  pivot_wider(names_from = scenario, values_from = biomass_change_percent) %>%
  mutate(diff = `With maturity decline` - Baseline) %>%
  group_by(species, time) %>%
  summarise(
    mean_diff = mean(diff),
    min_diff  = min(diff),
    max_diff  = max(diff),
    .groups = 'drop'
  )

p_diff <- ggplot(baseline_diff_summary,
                 aes(x = time, y = mean_diff, color = species, fill = species)) +
  geom_ribbon(aes(ymin = min_diff, ymax = max_diff), alpha = 0.2, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", alpha = 0.7) +
  labs(x = "Time (years)",
       y = "Difference from baseline (%)",
       color = "Species", fill = "Species") +
  scale_color_manual(values = species_colors) +
  scale_fill_manual(values = species_colors) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.line.x.bottom = element_line(color = "black", linewidth = 0.5),
    axis.line.y.left   = element_line(color = "black", linewidth = 0.5),
    axis.text   = element_text(size = 9, color = "black"),
    axis.title  = element_text(size = 11, face = "bold"),
    axis.ticks  = element_line(color = "black", linewidth = 0.5),
    legend.position   = "right",
    legend.title      = element_text(size = 10, face = "bold"),
    legend.text       = element_text(size = 9),
    legend.key        = element_blank(),
    plot.background = element_rect(fill = "white", color = NA),
    plot.margin     = margin(10, 10, 10, 10)
  )

print(p_diff)
```


```{r}
sp_order
# compare against the order panels actually appear:
sort(unique(as.character(abs_band_tonnes$sp)))


effort_long <- as.data.frame(f_main) %>%
  tibble::rownames_to_column("year") %>%
  mutate(year = as.numeric(year)) %>%
  tidyr::pivot_longer(-year, names_to = "sp", values_to = "effort")

scale_tbl <- abs_band_tonnes %>%
  group_by(sp) %>% summarise(ymax_panel = max(ymax, na.rm = TRUE), .groups = "drop") %>%
  left_join(
    effort_long %>% group_by(sp) %>% summarise(emax = max(effort, na.rm = TRUE), .groups = "drop"),
    by = "sp"
  ) %>%
  mutate(sf = ymax_panel / emax)

effort_long <- effort_long %>%
  left_join(scale_tbl, by = "sp") %>%
  mutate(effort_scaled = effort * sf)

library(ggh4x)
sp_order <- levels(factor(abs_band_tonnes$sp))   # panel order top-left → bottom-right

y_scales <- lapply(sp_order, function(s) {
  sf  <- scale_tbl$sf[scale_tbl$sp == s]
  emx <- scale_tbl$emax[scale_tbl$sp == s]
  brks <- pretty(c(0, emx), n = 3)
  brks <- brks[brks <= emx * 1.02]
  scale_y_continuous(
    limits   = c(0, NA),
    expand   = expansion(mult = c(0, 0.05)),
    sec.axis = sec_axis(~ . / sf, name = "Fishing mortality (F)", breaks = brks)
  )
})
```

```{r}
# ============================================================
# SETUP / CALCULATIONS
# ============================================================
library(ggh4x)

area_km2 <- 381.6

# --- Absolute yield band (ribbon data), converted to tonnes ---
abs_band <- y_realistic %>%
  group_by(sp, year, scenario) %>%
  summarise(ymin = min(value), ymax = max(value),
            ymean = mean(value), .groups = "drop")

abs_band_tonnes <- abs_band %>%
  mutate(
    ymin  = ymin  * area_km2,
    ymax  = ymax  * area_km2,
    ymean = ymean * area_km2
  )

# Drop unused factor levels (small_fish, ruffe have no panels here)
# so panels render only for species present, in factor-level order:
# bream, white_bream, roach, perch, pikeperch
abs_band_tonnes$sp <- droplevels(abs_band_tonnes$sp)

# --- Cumulative loss labels (from percent-change pipeline) ---
yield_changes <- y_realistic %>%
  select(step, year, sp, model, scenario, value) %>%
  pivot_wider(names_from = scenario, values_from = value) %>%
  mutate(percent_change = ((`With maturity decline` - Baseline) / Baseline) * 100)

uncertainty <- yield_changes %>%
  group_by(sp, step, year) %>%
  summarise(min_percent_change = min(percent_change, na.rm = TRUE),
            max_percent_change = max(percent_change, na.rm = TRUE),
            .groups = "drop")

cumulative_loss_labels <- uncertainty %>%
  group_by(sp) %>%
  summarise(total_min_loss = sum(min_percent_change, na.rm = TRUE),
            total_max_loss = sum(max_percent_change, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(label = paste0("Cumulative loss: ", round(total_max_loss, 0),
                        " — ", round(total_min_loss, 0), "%"),
         x_pos = decline_end - 1, y_pos = -1)

# --- Effort data (F values) reshaped long, names matched to yield ---
effort_long <- as.data.frame(f_main) %>%
  tibble::rownames_to_column("year") %>%
  mutate(year = as.numeric(year)) %>%
  tidyr::pivot_longer(-year, names_to = "sp", values_to = "effort")

# Align effort sp factor to the same levels/order as the yield data
effort_long$sp <- factor(effort_long$sp, levels = levels(abs_band_tonnes$sp))

# --- Per-species scale factor (yield max / effort max) ---
scale_tbl <- abs_band_tonnes %>%
  group_by(sp) %>% summarise(ymax_panel = max(ymax, na.rm = TRUE), .groups = "drop") %>%
  left_join(
    effort_long %>% group_by(sp) %>% summarise(emax = max(effort, na.rm = TRUE), .groups = "drop"),
    by = "sp"
  ) %>%
  mutate(sf = ymax_panel / emax)

# sanity check — should have NO NA / Inf in emax or sf
print(scale_tbl)

effort_long <- effort_long %>%
  left_join(scale_tbl, by = "sp") %>%
  mutate(effort_scaled = effort * sf)

# ============================================================
# PER-PANEL Y-SCALES
# ============================================================
# Panel render order = factor levels, restricted to species present
sp_order <- levels(abs_band_tonnes$sp)
sp_order <- sp_order[sp_order %in% unique(as.character(abs_band_tonnes$sp))]
# → bream, white_bream, roach, perch, pikeperch

y_scales <- lapply(sp_order, function(s) {
  row <- scale_tbl[scale_tbl$sp == s, ]
  sf  <- row$sf
  emx <- row$emax

  # No valid effort for this species → plain axis, no secondary axis
  if (length(sf) == 0 || is.na(sf) || !is.finite(sf) || sf <= 0) {
    return(scale_y_continuous(limits = c(0, NA),
                              expand = expansion(mult = c(0, 0.05))))
  }

  brks <- pretty(c(0, emx), n = 3)
  brks <- brks[brks > 0 & brks <= emx]          # valid interior ticks only
  if (length(brks) == 0) brks <- round(emx, 2)  # guard: never empty

  scale_y_continuous(
    limits   = c(0, NA),
    expand   = expansion(mult = c(0, 0.05)),
    sec.axis = sec_axis(~ . / sf, name = "Fishing mortality (F)",
                        breaks = c(0, brks))
  )
})

# ============================================================
# PLOT
# ============================================================
yield_absolute <- ggplot() +
  geom_ribbon(data = abs_band_tonnes,
              aes(x = year, ymin = ymin, ymax = ymax, fill = scenario),
              alpha = 0.7) +
  geom_vline(xintercept = -Inf, color = "gray5", linewidth = 0.3) +
  geom_hline(yintercept = -Inf, color = "gray5", linewidth = 0.3) +
  geom_line(data = effort_long,
            aes(x = year, y = effort_scaled),
            color = "gray5", linewidth = 0.3, inherit.aes = FALSE) +
  geom_text(data = cumulative_loss_labels,
            aes(x = x_pos, y = -Inf, label = label),
            size = 4, fontface = "bold", hjust = 0.95, vjust = -0.5,
            inherit.aes = FALSE) +
  facet_wrap2(~sp, scales = "free_y", ncol = 2,
              labeller = as_labeller(function(x) gsub("_", " ", x))) +
  facetted_pos_scales(y = y_scales) +
  scale_color_manual(values = c("Baseline" = "grey40",
                                "With maturity decline" = "firebrick"),
                     labels = c("Baseline" = "Baseline",
                                "With maturity decline" = "Declining size at maturity")) +
  scale_fill_manual(values  = c("Baseline" = "grey40",
                                "With maturity decline" = "firebrick"),
                    labels  = c("Baseline" = "Baseline",
                                "With maturity decline" = "Declining size at maturity")) +
  labs(x = "Time (year)", y = "Yield (tonnes)", color = NULL, fill = NULL) +
  My_theme +
  theme(
  legend.position    = "bottom",
  axis.line.y.right  = element_line(color = "gray5", linewidth = 0.3)
)

yield_absolute

# ggsave("yield_absoluteF.tiff", yield_absolute, width = 9, height = 8, dpi = 600, bg = "white")
```


```{r}
# Convert yields to tonnes by multiplying by area
area_km2 <- 381.6

# === PLOT A: ABSOLUTE YIELDS (wiggles visible) ===
abs_band <- y_realistic %>%
  group_by(sp, year, scenario) %>%
  summarise(ymin = min(value), ymax = max(value),
            ymean = mean(value), .groups = "drop")

abs_band_tonnes <- abs_band %>%
  mutate(
    ymin  = ymin  * area_km2,
    ymax  = ymax  * area_km2,
    ymean = ymean * area_km2
  )

# ggsave("yield_loss.tiff", yield_loss, width = 7, height = 7, dpi = 600, bg = "white")

yield_absolute <- ggplot() +
  geom_ribbon(data = abs_band_tonnes,
              aes(x = year, ymin = ymin, ymax = ymax, fill = scenario),
              alpha = 0.7) +
  geom_vline(xintercept = -Inf, color = "gray5", linewidth = 0.3) +
  geom_hline(yintercept = -Inf, color = "gray5", linewidth = 0.3) +
  geom_line(data = effort_long,
          aes(x = year, y = effort_scaled),
          color = "black", linewidth = 0.3, inherit.aes = FALSE) +
  #geom_line(data = abs_band_tonnes,
  #          aes(x = year, y = ymean, color = scenario),
  #          linewidth = 0.8) +
  geom_text(data = cumulative_loss_labels,
          aes(x = x_pos, y = -Inf, label = label),
          size = 4, fontface = "bold", hjust = 1, vjust = -0.5,
          inherit.aes = FALSE) +
  facet_wrap2(~sp, scales = "free_y", ncol = 2,
            labeller = as_labeller(function(x) gsub("_", " ", x))) +
  facetted_pos_scales(y = y_scales) +
  scale_color_manual(values = c("Baseline" = "grey40",
                                "With maturity decline" = "firebrick"),
                     labels = c("Baseline" = "Baseline",
                                "With maturity decline" = "Declining size at maturity")) +
  scale_fill_manual(values  = c("Baseline" = "grey40",
                                "With maturity decline" = "firebrick"),
                    labels  = c("Baseline" = "Baseline",
                                "With maturity decline" = "Declining size at maturity")) +
  labs(x = "Time (year)", y = "Yield (tonnes)", color = NULL, fill = NULL) +
  My_theme +
  theme(legend.position = "bottom")

# ggsave("yield_absolute.tiff", yield_absolute, width = 8, height = 8, dpi = 600, bg = "white")

# === PLOT B: PERCENT CHANGE (smooth maturity signal) ===
yield_changes <- y_realistic %>%
  select(step, year, sp, model, scenario, value) %>%
  pivot_wider(names_from = scenario, values_from = value) %>%
  mutate(percent_change = ((`With maturity decline` - Baseline) / Baseline) * 100)

uncertainty <- yield_changes %>%
  group_by(sp, step, year) %>%
  summarise(min_percent_change = min(percent_change, na.rm = TRUE),
            max_percent_change = max(percent_change, na.rm = TRUE),
            .groups = "drop")

cumulative_loss_labels <- uncertainty %>%
  group_by(sp) %>%
  summarise(total_min_loss = sum(min_percent_change, na.rm = TRUE),
            total_max_loss = sum(max_percent_change, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(label = paste0("Cumulative loss: ", round(total_max_loss, 0),
                        " — ", round(total_min_loss, 0), "%"),
         x_pos = decline_end - 1, y_pos = -1)


yield_loss <- ggplot() +
  geom_ribbon(data = uncertainty,
              aes(x = year, ymin = min_percent_change, ymax = max_percent_change),
              fill = "grey70", alpha = 0.4) +
  geom_line(data = yield_changes,
            aes(x = year, y = percent_change, group = model),
            color = "grey20", alpha = 0.6, linewidth = 0.7) +
  geom_hline(yintercept = 0,    linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = -Inf, color = "gray5", linewidth = 0.3) +
  geom_hline(yintercept = -Inf, color = "gray5", linewidth = 0.3) +
  geom_text(data = cumulative_loss_labels,
            aes(x = x_pos, y = y_pos, label = label),
            size = 4, fontface = "bold", hjust = 1, vjust = 1, inherit.aes = FALSE) +
  facet_wrap(~sp, scales = "fixed", ncol = 2,
            labeller = as_labeller(function(x) gsub("_", " ", x))) +
  labs(x = "Time (year)", y = "Percent change (%)") +
  My_theme

# ggsave("yield_loss.tiff", yield_loss, width = 7, height = 7, dpi = 600, bg = "white")

# write.csv(abs_band, "abs_band.csv", row.names = FALSE)
```


-------------THE END---------- 🐟🎣
